---
title: "Protrudin_spine_type_analysis"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r spine_type_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = 'png', message = FALSE)
knitr::opts_chunk$set(fig.path = "../figures/")

source('../style/thesis_setup.R')

library(stringr)
library(reshape2)
#library(dplyr) calling explicitly
library(ggplot2)
library(rstatix)
library(ggpubr)
library(here)
library(readbitmap)
library(patchwork)
```


CSV files with spine counts for each experiment were generated from the ImageJ Cell Counter output files (in the 'cell_counter' folder for each experiment) using the code below
```{r embedded_Cell_counter_XMLs_to_CSV, eval=FALSE}
library(xml2)

dir = choose.dir(caption = "Choose directory containing Cell Counter XML files.")
files = list.files(dir, pattern = '.(xml|XML)$', recursive = F)


counts = do.call(rbind, lapply(files, function(f){
  xml = read_xml(paste(dir, f, sep = '/'))
  xml.marker_data = xml_child(xml, search = 'Marker_Data')
  
  mushroom = xml_child(xml.marker_data, search = 2)
  thin = xml_child(xml.marker_data, search = 3)
  stubby = xml_child(xml.marker_data, search = 4)
  filopodia = xml_child(xml.marker_data, search = 5)
  
  mushroom.count = xml_length(mushroom) - 2 #subtracting to account for type and name fields
  thin.count = xml_length(thin) - 2
  stubby.count = xml_length(stubby) - 2
  filopodia.count = xml_length(filopodia) - 2
  
  df = data.frame(file = f, 
                  mushroom = mushroom.count,
                  thin = thin.count,
                  stubby = stubby.count,
                  filopodia = filopodia.count)
  return(df)
  
}))

write.csv(counts, file = paste(dir, "cell_counter_counts.csv", sep = '/'))
```

Reading in CSV files
```{r read_CSVs}
CNC78 = read.csv(here("data/spine_classification/CNC78/cell_counter/cell_counter_counts.csv"), stringsAsFactors = F)
CNC79 = read.csv(here("data/spine_classification/CNC79/cell_counter/cell_counter_counts.csv"), stringsAsFactors = F)
CNC80 = read.csv(here("data/spine_classification/CNC80/cell_counter/cell_counter_counts.csv"), stringsAsFactors = F)
CNC79_DIV21 = read.csv(here("data/spine_classification/CNC79_DIV21/cell_counter/cell_counter_counts.csv"), stringsAsFactors = F)
```

Adding columns to each table for experiment number and cell age
```{r expt_and_age}
CNC78$expt = rep("78", nrow(CNC78))
CNC78$age = rep("DIV15", nrow(CNC78))

CNC79$expt = rep("79", nrow(CNC79))
CNC79$age = rep("DIV15", nrow(CNC79))

CNC80$expt = rep("80", nrow(CNC80))
CNC80$age = rep("DIV15", nrow(CNC80))

CNC79_DIV21$expt = rep("79", nrow(CNC79_DIV21))
CNC79_DIV21$age = rep("DIV21", nrow(CNC79_DIV21))
```

Pulling condition (i.e. protrudin mutant) and cell number from the file names
```{r cond_and_cell}
CNC78$cond = str_match(CNC78$file, "CellCounter_(.+)_(C\\d+)")[,2]
CNC78$cell = str_match(CNC78$file, "CellCounter_(.+)_(C\\d+)")[,3]

CNC79$cond = str_match(CNC79$file, "CellCounter_(.+)_(C\\d+)")[,2]
CNC79$cell = str_match(CNC79$file, "CellCounter_(.+)_(C\\d+)")[,3]

CNC80$cond = str_match(CNC80$file, "CellCounter_(.+)_(C\\d+)")[,2]
CNC80$cell = str_match(CNC80$file, "CellCounter_(.+)_(C\\d+)")[,3]

CNC79_DIV21$cond = str_match(CNC79_DIV21$file, "CellCounter_(.+)_(C\\d+)")[,2]
CNC79_DIV21$cell = str_match(CNC79_DIV21$file, "CellCounter_(.+)_(C\\d+)")[,3]
```

Combining into a single table, adding total and normalised spine counts for each cell, and creating 'melted' versions to use for certain plots
```{r full_data_frames}
spine_counts = rbind(CNC78, CNC79, CNC80, CNC79_DIV21)
spine_counts$total = sapply(1:nrow(spine_counts), function(i){ sum(spine_counts$mushroom[i], spine_counts$thin[i], spine_counts$stubby[i], spine_counts$filopodia[i]) })
spine_counts$m_norm = spine_counts$mushroom/spine_counts$total
spine_counts$t_norm = spine_counts$thin/spine_counts$total
spine_counts$s_norm = spine_counts$stubby/spine_counts$total
spine_counts$f_norm = spine_counts$filopodia/spine_counts$total
spine_counts$cell_id = paste0(spine_counts$expt, spine_counts$age, spine_counts$cond, spine_counts$cell)

spine_counts.normalised.melt = melt(spine_counts, id.vars = c('expt', 'age', 'cond', 'cell_id'), measure.vars = c('m_norm', 't_norm', 's_norm', 'f_norm'))
spine_counts.melt = melt(spine_counts, id.vars = c('expt', 'age', 'cond', 'cell_id'), measure.vars = c('mushroom', 'thin', 'stubby', 'filopodia'))

spine_counts.melt = spine_counts.melt %>% dplyr::rename(spine_type = variable)
spine_counts.normalised.melt = spine_counts.normalised.melt %>% dplyr::rename(spine_type = variable)

#converting condition columns to factors for plot ordering and labels
spine_counts$cond = factor(spine_counts$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)
spine_counts.normalised.melt$cond = factor(spine_counts.normalised.melt$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)
spine_counts.melt$cond = factor(spine_counts.melt$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)



spine_counts.normalised.melt$spine_type = factor(spine_counts.normalised.melt$spine_type, levels = c('m_norm', 't_norm', 's_norm', 'f_norm'), labels = c('mushroom', 'thin', 'stubby', 'filopodia'))

```

# Plots

## DIV15

I transfected cortical neurons at DIV10 and imaged dendritic spines at DIV15 in three experiments so far (CNC #78, 79, and 80).  Each plate of cells was transfected with 1μg of EGFP/Emerald, as well as 2μg of one of the following red SYN constructs: mCherry, WT protrudin, Mut2 protrudin, ΔEM, ΔFFAT, ΔRBD, ΔKIF5, or ΔFYVE.  (CNC #78 did not include ΔKIF5 or ΔFYVE because there were not enough plates.)  In each case, I imaged 10 cells for each condition (on the spinning disc with 100x objective, as a Z_stack through the soma and an area of the dendrites), and for each cell selected one, 20μm section of dendrite to count spines.  

### Total number of spines
Stats with respect to mCherry control...
```{r DIV15_total_mCh_stat}
DIV15.counts.dunn = dunn_test(spine_counts[spine_counts$age == 'DIV15',], total ~ cond, p.adjust.method = 'bonferroni')
DIV15.counts.dunn = add_xy_position(DIV15.counts.dunn, x = 'cond')
DIV15.counts.dunn = add_y_position(DIV15.counts.dunn, step.increase = 0.01)
DIV15.counts.dunn$p.adj.rounded = sapply(DIV15.counts.dunn$p.adj, function(p){signif(p, digits = 3)})


DIV15_count = ggplot(spine_counts[spine_counts$age == 'DIV15',], aes(x = cond, y = total)) + 
  geom_dotplot(aes(fill = cond, colour = cond), binaxis = "y", stackdir = "center", position = "dodge", binwidth = 1, dotsize = 0.5) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  stat_compare_means(label.x = 2) + stat_pvalue_manual(DIV15.counts.dunn, label = 'p.adj.rounded', hide.ns = T, position = position_nudge(y = -1)) + 
  labs(x = '', y = 'Number of spines in 20 μm section', title = 'DIV15') +
  ylim(-1, 30) + 
  d_format + report_theme

DIV15_count
```
...and the same again but with WT protrudin as the reference group
```{r DIV15_total_WT_stat, fig.width=8, fig.height=6}
ggplot(spine_counts[spine_counts$age == 'DIV15',], aes(x = cond, y = total, group = cond, fill = cond, colour = cond)) + 
  geom_dotplot(binaxis = "y", stackdir = "center", position = "dodge", dotsize = 0.7) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  stat_compare_means(label = 'p.signif', ref.group = 'WT', hide.ns = F) + 
  labs(x = 'Condition', y = 'Number of spines in 20 μm section', title = 'Dendritic spines at DIV15') +
  d_format + report_theme
```



### Breakdown by spine classification
Overall number of each type of spine, stats w.r.t. mCherry
```{r DIV15_type_absolute, fig.width=8, fig.height=6}
ggplot(spine_counts.melt[spine_counts.melt$age == 'DIV15',], aes(x = cond, y = value, group = cond, fill = cond, colour = cond)) + 
  geom_dotplot(binaxis = "y", stackdir = "center", position = "dodge", dotsize = 0.7) + geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  facet_wrap(~spine_type) + 
  stat_compare_means(label = 'p.signif', ref.group = 'Control') + 
  labs(x = 'Condition', y = 'Number of spines in 20 μm section', title = 'Dendritic spines at DIV15 by type') +
  d_format + report_theme

```

Normalised to total number of spines on each cell
```{r DIV15_type_norm, fig.width = 8, fig.height=10}
DIV15.test <- spine_counts.normalised.melt[spine_counts.normalised.melt$age == 'DIV15',] %>% group_by(spine_type) %>% dplyr::select(value, cond)
DIV15.dunn = dunn_test(DIV15.test, value ~ cond, p.adjust.method = 'bonferroni')
DIV15.dunn = add_xy_position(DIV15.dunn, x = 'cond')
DIV15.dunn = add_y_position(DIV15.dunn, step.increase = 0.04)
DIV15.dunn$p.adj.rounded = sapply(DIV15.dunn$p.adj, function(p){signif(p, digits = 3)})

DIV15_type_norm = ggplot(spine_counts.normalised.melt[spine_counts.normalised.melt$age == 'DIV15',], aes(x = cond, y = value)) + 
  geom_dotplot(aes(fill = cond, colour = cond), binaxis = "y", stackdir = "center", position = "dodge", dotsize = 2, binwidth = 0.025) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  facet_wrap(~spine_type) + 
  scale_y_continuous(limits = c(-0.05, 1.4), breaks = c(0.0, 0.5, 1.0)) + 
  stat_compare_means(label.x = 1.5, label.y = 1.35) + stat_pvalue_manual(DIV15.dunn, hide.ns = TRUE, label = 'p.adj.rounded', position = position_nudge(y = -0.15)) + 
  labs(x = '', y = 'Number of spines/total spines in 20 μm section', title = 'DIV15') +
  d_format + report_theme

DIV15_type_norm
```

Another way of looking at percentages
```{r bar_fill_DIV15, fig.width=8, fig.height=6}
ggplot(spine_counts.melt[spine_counts.melt$age == 'DIV15',], aes(x = cell_id, y = value, fill = spine_type)) + 
  geom_bar(position="fill", stat="identity") + 
  facet_wrap(~cond, scales = "free_x") + 
  labs(x = '', y = 'Spine type proportion', title = 'Spine type proportions at DIV15') + 
  d_format + report_theme
```
(The gap in the FYVE box is a cell without any spines.)

So it looks like maybe WT is doing something, but the others aren't.  If protrudin has some sort of a role in spine pruning or turnover, maybe the mutations interfere with that function.  I'm not sure what the story is with Mutant 2 though.  Maybe this is something that protrudin does that isn't directly linked to how it promotes regeneration?

## DIV21
CNC #79 was left to grow to DIV21 and imaged again.  This is only one experiment so far (10 cells per condition), so it's very preliminary data.

### Total number of spines
Stats with respect to mCherry control...
```{r DIV21_total_mCh_stat}
DIV21.counts.dunn = dunn_test(spine_counts[spine_counts$age == 'DIV21',], total ~ cond, p.adjust.method = 'bonferroni')
DIV21.counts.dunn = add_xy_position(DIV21.counts.dunn, x = 'cond')
DIV21.counts.dunn = add_y_position(DIV21.counts.dunn, step.increase = 0)
DIV21.counts.dunn$p.adj.rounded = sapply(DIV21.counts.dunn$p.adj, function(p){signif(p, digits = 3)})
#individual comparisons non-significant

DIV21_count = ggplot(spine_counts[spine_counts$age == 'DIV21',], aes(x = cond, y = total)) + 
  geom_dotplot(aes(fill = cond, colour = cond), binaxis = "y", stackdir = "center", position = "dodge", binwidth = 1, dotsize = 0.5) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  stat_compare_means(label.x = 2) + stat_pvalue_manual(DIV21.counts.dunn, label = 'p.adj.rounded', hide.ns = T) + 
  labs(x = '', y = 'Number of spines in 20 μm section', title = 'DIV21') +
  ylim(-1, 30) + 
  d_format + report_theme

DIV21_count
```
...and the same again but with WT protrudin as the reference group
```{r DIV21_total_WT_stat, fig.width=8, fig.height=6}
ggplot(spine_counts[spine_counts$age == 'DIV21',], aes(x = cond, y = total, group = cond, fill = cond, colour = cond)) + 
  geom_dotplot(binaxis = "y", stackdir = "center", position = "dodge", dotsize = 0.7) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  stat_compare_means(label = 'p.signif', ref.group = 'WT', hide.ns = F) + 
  labs(x = 'Condition', y = 'Number of spines in 20 μm section', title = 'Dendritic spines at DIV21') +
  d_format + report_theme
```

### Breakdown by spine classification
Overall number of each type of spine, stats w.r.t. mCherry
```{r DIV21_type_absolute, fig.width=8, fig.height=6}
ggplot(spine_counts.melt[spine_counts.melt$age == 'DIV21',], aes(x = cond, y = value, group = cond, fill = cond, colour = cond)) + 
  geom_dotplot(binaxis = "y", stackdir = "center", position = "dodge", dotsize = 0.7) + geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  facet_wrap(~spine_type) + 
  stat_compare_means(label = 'p.signif', ref.group = 'Control') + 
  labs(x = 'Condition', y = 'Number of spines in 20 μm section', title = 'Dendritic spines at DIV21 by type') +
  d_format + report_theme

```

Normalised to total number of spines on each cell
```{r DIV21_type_norm, fig.width=8, fig.height = 10}

DIV21.test <- spine_counts.normalised.melt[spine_counts.normalised.melt$age == 'DIV21',] %>% group_by(spine_type) %>% dplyr::select(value, cond)
DIV21.dunn = dunn_test(DIV21.test, value ~ cond, p.adjust.method = 'bonferroni')
DIV21.dunn = add_xy_position(DIV21.dunn, x = 'cond')
DIV21.dunn = add_y_position(DIV21.dunn, step.increase = 0)
DIV21.dunn$p.adj.rounded = sapply(DIV21.dunn$p.adj, function(p){signif(p, digits = 3)})
#non-significant

DIV21_type_norm = ggplot(spine_counts.normalised.melt[spine_counts.normalised.melt$age == 'DIV21',], aes(x = cond, y = value)) + 
  geom_dotplot(aes(fill = cond, colour = cond), binaxis = "y", stackdir = "center", position = "dodge", dotsize = 2, binwidth = 0.025) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  facet_wrap(~spine_type) + 
  scale_y_continuous(limits = c(-0.05, 1.4), breaks = c(0.0, 0.5, 1.0)) + 
  stat_compare_means(label.x = 1.5, label.y = 1.35) +  stat_pvalue_manual(DIV21.dunn, label = 'p.adj.rounded', hide.ns = T) + 
  labs(x = '', y = 'Number of spines/total spines in 20 μm section', title = 'DIV21') +
  d_format + report_theme

DIV21_type_norm
```

Another way of looking at percentages
```{r bar_fill_DIV21, fig.width=8, fig.height=6}
ggplot(spine_counts.melt[spine_counts.melt$age == 'DIV21',], aes(x = cell_id, y = value, fill = spine_type)) + 
  geom_bar(position="fill", stat="identity") + 
  facet_wrap(~cond, scales = "free_x") + 
  labs(x = '', y = 'Spine type proportion', title = 'Spine type proportions at DIV21') + 
  d_format + report_theme
```

Maybe Mutant 2 has fewer filopodia at this stage, but it's too early to be sure.



```{r spine_types_full_figure, fig.width = 16, fig.height = 22}
schematic = image_as_ggplot(read.bitmap(here('figures/external/dendrite_section_schematic.png')))


spine_type_layout = 'AABC
                    AABC
                    DDDD
                    DDDD
                    DDDD
                    EEEE
                    EEEE
                    EEEE'

spine_types_plot = schematic + DIV15_count + DIV21_count + DIV15_type_norm + DIV21_type_norm + plot_layout(design = spine_type_layout) + plot_annotation(tag_levels = 'A') & theme(axis.text.x = element_text(angle = 45, hjust = 1))

#ggsave('test.png', plot = spine_types_plot, device = 'png') - FINE - issue is with latex

spine_types_plot
```