---
title: "Protrudin_spine_length_analysis"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dev = 'png')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
knitr::opts_chunk$set(fig.path = "../figures/")

source('../style/thesis_setup.R')
```

# Setup and data handling

Loading R packages used in this file
```{r packages}
library(stringr)
library(reshape2)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(patchwork)
```

Reading in CSV files
```{r read_spine_type_CSVs}
#CNC79 - copied and pasted for 80 and 78
CNC79.dir = 'C:/Users/fl299/PhD/Data/Mine/Spinning\ Disc/070721_cnc79_dendrites/max_projections/GFP/dendrite_sections/spines/'
CNC79.csvs = lapply(list.files(path = CNC79.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC79.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine lngth CSVs
#WARNING - reassigned for each experiment - start looking here if any bugs pop up
colnames = unique(lapply(CNC79.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC79.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC79.csvs = CNC79.csvs[which(lapply(lapply(CNC79.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC79 = do.call(rbind, CNC79.csvs)

CNC79$expt = rep("79", nrow(CNC79))
CNC79$age = rep("DIV15", nrow(CNC79))

#cleanup
remove(CNC79.dir, CNC79.csvs, colnames, colnames.counts, colnames.correct)


#CNC80
CNC80.dir = 'C:/Users/fl299/PhD/Data/Mine/Spinning\ Disc/140721_CNC80_dendrites/max_projections/GFP/dendrite_sections/spines/'
CNC80.csvs = lapply(list.files(path = CNC80.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC80.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine lngth CSVs
colnames = unique(lapply(CNC80.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC80.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC80.csvs = CNC80.csvs[which(lapply(lapply(CNC80.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC80 = do.call(rbind, CNC80.csvs)

CNC80$expt = rep("80", nrow(CNC80))
CNC80$age = rep("DIV15", nrow(CNC80))

#cleanup
remove(CNC80.dir, CNC80.csvs, colnames, colnames.counts, colnames.correct)


#CNC78
CNC78.dir = 'C:/Users/fl299/PhD/Data/Mine/Spinning\ Disc/300621_CNC78_dendrites/max_projections/GFP/dendrite_sections/spines/'
CNC78.csvs = lapply(list.files(path = CNC78.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC78.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine length CSVs
colnames = unique(lapply(CNC78.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC78.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC78.csvs = CNC78.csvs[which(lapply(lapply(CNC78.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC78 = do.call(rbind, CNC78.csvs)

CNC78$expt = rep("78", nrow(CNC78))
CNC78$age = rep("DIV15", nrow(CNC78))

#cleanup
remove(CNC78.dir, CNC78.csvs, colnames, colnames.counts, colnames.correct)

```
Combine experiments into single data frame
```{r combo_df}
spine_lengths.df = do.call(rbind, list(CNC79, CNC80, CNC78))

```


Parsing filenames for experimental condition and cell number
```{r parse_filenames}
spine_lengths.df$id = str_match(spine_lengths.df$file, "(.+)_(C\\d+)")[,1]
spine_lengths.df$cond = str_match(spine_lengths.df$file, "(.+)_(C\\d+)")[,2]
spine_lengths.df$cell = str_match(spine_lengths.df$file, "(.+)_(C\\d+)")[,3]

spine_lengths.df$id = paste(spine_lengths.df$id, spine_lengths.df$expt, spine_lengths.df$age, sep = '_')

```

Traced spine counts
```{r spine_counts_from_trace}
spine_counts.df = as.data.frame(table(spine_lengths.df$id), stringsAsFactors = F)
spine_counts.df$cond = str_match(spine_counts.df$Var1, "(.+)_(C\\d+)")[,2]
spine_counts.df$cell = str_match(spine_counts.df$Var1, "(.+)_(C\\d+)")[,3]

spine_counts.df$cond = factor(spine_counts.df$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)

```

```{r spine_counts_plot}
#dunn test for p values
spinningdisc.dunn = dunn_test(spine_counts.df, Freq ~ cond, p.adjust.method = 'bonferroni')
spinningdisc.dunn = add_xy_position(spinningdisc.dunn, x = "cond")
spinningdisc.dunn = add_y_position(spinningdisc.dunn, step.increase = 0.01)
spinningdisc.dunn$p.adj.rounded = sapply(spinningdisc.dunn$p.adj, function(p){round(p, digits = 3)})

counts.plot = ggplot(spine_counts.df, aes(x = cond, y = Freq)) + 
  geom_dotplot(aes(fill = cond, colour = cond), binaxis = "y", stackdir = "center", position = "dodge", dotsize = 0.7, binwidth = 1) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + 
  labs(title = 'Replicate 1', x = '', y = paste0('Number of spines in 20', '\U03BC', 'm section')) + 
  d_format + report_theme +
  stat_compare_means() + stat_pvalue_manual(spinningdisc.dunn, hide.ns = T, label = 'p.adj.rounded', step.increase = 0.01)

counts.plot
```

Spine length plots
```{r spine_lengths}
spine_lengths.melt = melt(spine_lengths.df, id.vars = c('expt', 'age', 'cond', 'cell', 'id'), measure.vars = 'Length')
spine_lengths.melt$cond = factor(spine_lengths.melt$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)

```

```{r spine_length_plots}
#dunn test for p values
lengths.dunn = dunn_test(spine_lengths.melt, value ~ cond, p.adjust.method = 'bonferroni')
lengths.dunn = add_xy_position(lengths.dunn, x = "cond")
lengths.dunn = add_y_position(lengths.dunn, step.increase = 0.1)
lengths.dunn$p.adj.rounded = sapply(lengths.dunn$p.adj, function(p){signif(p, digits = 3)})#signif instead of round for small p values


lengths.plot = ggplot(spine_lengths.melt, aes(x = cond, y = value, colour = cond)) + 
  geom_violin() + geom_boxplot(fill = NA, width = 0.2, outlier.shape = NA)  + 
  labs(title = 'Replicate 1', x = '', y = paste0('Spine length (', '\U03BC', 'm)')) + 
  d_format + report_theme +
  stat_compare_means() + stat_pvalue_manual(lengths.dunn, hide.ns = TRUE, label = 'p.adj.rounded', bracket.nudge.y = -2)

lengths.plot
```




# Confocal spines
```{r read_CSVs_confocal}
#coped and pasted from CNC79

CNC81.dir = 'C:\\Users\\fl299\\PhD\\Data\\Mine\\Confocal\\CNC81_dendrites\\GFP_max_projections\\dendrite_sections\\spines\\'
CNC81.csvs = lapply(list.files(path = CNC81.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC81.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine lngth CSVs
#WARNING - reassigned for each experiment - start looking here if any bugs pop up
colnames = unique(lapply(CNC81.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC81.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC81.csvs = CNC81.csvs[which(lapply(lapply(CNC81.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC81 = do.call(rbind, CNC81.csvs)

CNC81$expt = rep("81", nrow(CNC81))
CNC81$age = rep("DIV15", nrow(CNC81))

#cleanup
remove(CNC81.dir, CNC81.csvs, colnames, colnames.counts, colnames.correct)






CNC82.dir = 'C:\\Users\\fl299\\PhD\\Data\\Mine\\Confocal\\CNC82_dendrites\\GFP_max_projections\\dendrite_sections\\spines\\'
CNC82.csvs = lapply(list.files(path = CNC82.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC82.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine lngth CSVs
#WARNING - reassigned for each experiment - start looking here if any bugs pop up
colnames = unique(lapply(CNC82.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC82.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC82.csvs = CNC82.csvs[which(lapply(lapply(CNC82.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC82 = do.call(rbind, CNC82.csvs)

CNC82$expt = rep("82", nrow(CNC82))
CNC82$age = rep("DIV15", nrow(CNC82))

#cleanup
remove(CNC82.dir, CNC82.csvs, colnames, colnames.counts, colnames.correct)






CNC83.dir = 'C:\\Users\\fl299\\PhD\\Data\\Mine\\Confocal\\CNC83_dendrites\\GFP_max_projections\\dendrite_sections\\spines\\'
CNC83.csvs = lapply(list.files(path = CNC83.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC83.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine lngth CSVs
#WARNING - reassigned for each experiment - start looking here if any bugs pop up
colnames = unique(lapply(CNC83.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC83.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC83.csvs = CNC83.csvs[which(lapply(lapply(CNC83.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC83 = do.call(rbind, CNC83.csvs)

CNC83$expt = rep("83", nrow(CNC83))
CNC83$age = rep("DIV15", nrow(CNC83))

#cleanup
remove(CNC83.dir, CNC83.csvs, colnames, colnames.counts, colnames.correct)






CNC84.dir = 'C:\\Users\\fl299\\PhD\\Data\\Mine\\Confocal\\CNC84_dendrites\\GFP_max_projections\\dendrite_sections\\spines\\'
CNC84.csvs = lapply(list.files(path = CNC84.dir, pattern = '\\.csv$'), 
                              function(f){
                                df = read.csv(paste0(CNC84.dir, f), stringsAsFactors = F);
                                df$file = rep(f, nrow(df));
                                return(df);
                              })
#remove key.csvs if present - assuming that these will be fewer than desired spine lngth CSVs
#WARNING - reassigned for each experiment - start looking here if any bugs pop up
colnames = unique(lapply(CNC84.csvs, colnames))
colnames.counts = sapply(colnames, function(c){ sum(sapply(lapply(CNC84.csvs, colnames), function(r){length(r) == length(c) && r == c})) })
colnames.correct = colnames[[which(colnames.counts == max(colnames.counts))]]

CNC84.csvs = CNC84.csvs[which(lapply(lapply(CNC84.csvs, colnames), function(c){ all(c == colnames.correct) }) == TRUE)]

CNC84 = do.call(rbind, CNC84.csvs)

CNC84$expt = rep("84", nrow(CNC84))
CNC84$age = rep("DIV15", nrow(CNC84))

#cleanup
remove(CNC84.dir, CNC84.csvs, colnames, colnames.counts, colnames.correct)
```
Combine experiments into single data frame
```{r combo_df_confocal}
spine_lengths_confocal.df = do.call(rbind, list(CNC81, CNC82, CNC83, CNC84))

```


Parsing filenames for experimental condition and cell number
```{r parse_filenames_confocal}
spine_lengths_confocal.df$id = str_match(spine_lengths_confocal.df$file, "(.+)_(C\\d+)")[,1]
spine_lengths_confocal.df$cond = str_match(spine_lengths_confocal.df$file, "(.+)_(C\\d+)")[,2]
spine_lengths_confocal.df$cell = str_match(spine_lengths_confocal.df$file, "(.+)_(C\\d+)")[,3]

spine_lengths_confocal.df$id = paste(spine_lengths_confocal.df$id, spine_lengths_confocal.df$expt, spine_lengths_confocal.df$age, sep = '_')

```

Traced spine counts
```{r spine_counts_confocal_from_trace_confocal}
spine_counts_confocal.df = as.data.frame(table(spine_lengths_confocal.df$id), stringsAsFactors = F)
spine_counts_confocal.df$cond = str_match(spine_counts_confocal.df$Var1, "(.+)_(C\\d+)")[,2]
spine_counts_confocal.df$cell = str_match(spine_counts_confocal.df$Var1, "(.+)_(C\\d+)")[,3]

spine_counts_confocal.df$cond = factor(spine_counts_confocal.df$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)

```

```{r spine_counts_confocal_plot}
#dunn test for p values
confocal.dunn = dunn_test(spine_counts_confocal.df, Freq ~ cond, p.adjust.method = 'bonferroni')
confocal.dunn = add_xy_position(confocal.dunn, x = "cond")
confocal.dunn$p.adj.rounded = sapply(confocal.dunn$p.adj, function(p){round(p, digits = 3)})
#result - no individual comparisons are significant

counts_confocal.plot= ggplot(spine_counts_confocal.df, aes(x = cond, y = Freq)) + 
  geom_dotplot(aes(fill = cond, colour = cond), binaxis = "y", stackdir = "center", position = "dodge", dotsize = 1, binwidth = 1) + 
  geom_boxplot(fill = NA, colour = 'black', outlier.colour = NA) + labs(x = '', y = paste0('Number of spines in 20', '\U03BC', 'm section'), title = 'Replicate 2') + 
  d_format + report_theme +
  stat_compare_means() + stat_pvalue_manual(confocal.dunn, hide.ns = T, label = 'p.adj.rounded', step.increase = 0.1)

counts_confocal.plot
```


Spine length plots
```{r spine_lengths_confocal}
spine_lengths_confocal.melt = melt(spine_lengths_confocal.df, id.vars = c('expt', 'age', 'cond', 'cell', 'id'), measure.vars = 'Length')
spine_lengths_confocal.melt$cond = factor(spine_lengths_confocal.melt$cond, levels = c('mCh', 'WT', 'Mut2', 'ER', 'FFAT', 'RBD', 'KIF5', 'FYVE'), labels = c('Control', 'WT', 'Active', paste0('\U0394', 'ER'), paste0('\U0394', 'FFAT'), paste0('\U0394', 'RBD'), paste0('\U0394', 'KIF5'), paste0('\U0394', 'FYVE')), ordered = T)

```

```{r spine_length_plots_confocal}
#dunn test for p values
lengths_confocal.dunn = dunn_test(spine_lengths_confocal.melt, value ~ cond, p.adjust.method = 'bonferroni')
lengths_confocal.dunn = add_xy_position(lengths_confocal.dunn, x = "cond")
lengths_confocal.dunn = add_y_position(lengths_confocal.dunn, step.increase = 0.02)
lengths_confocal.dunn$p.adj.rounded = sapply(lengths_confocal.dunn$p.adj, function(p){round(p, digits = 3)})

lengths_confocal.plot = ggplot(spine_lengths_confocal.melt, aes(x = cond, y = value, colour = cond)) + 
  geom_violin() + geom_boxplot(fill = NA, width = 0.2, outlier.shape = NA) + 
  labs(title = 'Replicate 2', x = '', y = paste0('Spine length (', '\U03BC', 'm)')) + 
  d_format + report_theme +
  stat_compare_means() + stat_pvalue_manual(lengths_confocal.dunn, hide.ns = TRUE, label = 'p.adj.rounded')

lengths_confocal.plot
```

# Combined plots

```{r spine_lengths_full_figure, fig.height=14, fig.width=16, fig.path = '../figures/full/', dev = 'png'}
(counts.plot + counts_confocal.plot) / (lengths.plot + lengths_confocal.plot) + plot_annotation(tag_levels = 'A') & theme(axis.text.x = element_text(angle = 45, hjust=1))

```



