# Setup
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
dir.create("./figures/")
knitr::opts_chunk$set(fig.path = "figures/")


source('C:\\Users\\fl299\\GitHub\\kymograph-analysis\\Kymograph_trace_analysis.R')
source('C:\\Users\\fl299\\GitHub\\first-year-report\\data_analysis\\report_setup.R')
# GFP #8AC926
# WT #19647E
# MUT #FF595E
d_format = list(scale_colour_manual(values = c('control' = '#8AC926', 'WT' = '#19647E', 'Active' = '#FF595E')),
                 scale_fill_manual(values = c('control' = '#8AC926', 'WT' = '#19647E', 'Active' = '#FF595E')))
#from palette - https://coolors.co/1f2041-4b3f72-ffc857-1982c4-119da4-19647e-ff595e-8ac926

library(dplyr)
library(ggplot2)
library(gridExtra)
#library(plyr) #calling explicitly to avoid conflict with dplyr
library(ggpubr)
library(reshape2)
library(PMCMRplus)

library(readbitmap)
library(patchwork)
```

# Data handling
```{r csv_selection}

CNC57 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC57"
CNC58_59 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC58-59"
CNC59 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC59"
CNC61 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC61"

#keeping experiments separate because of frame rate differences


mCherry_CSVs_58 = list.files(CNC58_59, pattern = "^MCR_.+\\.csv$", full.names = TRUE)
WT_CSVs_58 = list.files(CNC58_59, pattern = "^WTR_.+\\.csv$", full.names = TRUE)
MUT2_CSVs_58 = list.files(CNC58_59, pattern = "^MR_.+\\.csv$", full.names = TRUE)


mCherry_CSVs_59 = list.files(CNC59, pattern = "^MCR_.+\\.csv$", full.names = TRUE)
WT_CSVs_59 = list.files(CNC59, pattern = "^WTR_.+\\.csv$", full.names = TRUE)
MUT2_CSVs_59 = list.files(CNC59, pattern = "^MR_.+\\.csv$", full.names = TRUE)

mCherry_CSVs_61 = list.files(CNC61, pattern = "^MCR_.+\\.csv$", full.names = TRUE)
WT_CSVs_61 = list.files(CNC61, pattern = "^WTR_.+\\.csv$", full.names = TRUE)
MUT2_CSVs_61 = list.files(CNC61, pattern = "^M2R_.+\\.csv$", full.names = TRUE)


```


```{r analyse_traces}
#TODO - fix frame rates to CSV for each stack


mCherry_data_58 = lapply(mCherry_CSVs_58, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC58"
  return(traces)
})
WT_data_58 = lapply(WT_CSVs_58, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC58"
  return(traces)
})
MUT2_data_58 = lapply(MUT2_CSVs_58, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC58"
  return(traces)
})

mCherry_data_59 = lapply(mCherry_CSVs_59, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC59"
  return(traces)
})
WT_data_59 = lapply(WT_CSVs_59, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC59"
  return(traces)
})
MUT2_data_59 = lapply(MUT2_CSVs_59, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC59"
  return(traces)
})

mCherry_data_61 = lapply(mCherry_CSVs_61, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC61"
  return(traces)
})
WT_data_61 = lapply(WT_CSVs_61, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC61"
  return(traces)
})
MUT2_data_61 = lapply(MUT2_CSVs_61, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC61"
  return(traces)
})





#combine


mCherry_Rab7_data = c(mCherry_data_58, mCherry_data_59, mCherry_data_61)
WT_Rab7_data = c(WT_data_58, WT_data_59, WT_data_61)
MUT2_Rab7_data = c(MUT2_data_58, MUT2_data_59, MUT2_data_61)



```


Basic analysis
```{r}
#all one video per cell in this case - don't need separate count and can number sequentially - LEAVING ARL8 FOR NOW
n_cells = list(mCherry = length(mCherry_Rab7_data), WT = length(WT_Rab7_data), MUT2 = length(MUT2_Rab7_data))

#pulling out 'per segment' data from analysed CSVs
mCherry.per_seg = do.call(rbind, lapply(1:length(mCherry_Rab7_data), function(c){ df = mCherry_Rab7_data[[c]]$per_seg; df$cell = c; df$cond = "mCherry"; df$exp = mCherry_Rab7_data[[c]]$experiment; return(df) }))
WT.per_seg = do.call(rbind, lapply(1:length(WT_Rab7_data), function(c){ df = WT_Rab7_data[[c]]$per_seg; df$cell = c; df$cond = "WT"; df$exp = WT_Rab7_data[[c]]$experiment; return(df) }))
MUT2.per_seg = do.call(rbind, lapply(1:length(MUT2_Rab7_data), function(c){ df = MUT2_Rab7_data[[c]]$per_seg; df$cell = c; df$cond = "MUT2"; df$exp = MUT2_Rab7_data[[c]]$experiment; return(df) }))


#combining data from all conditions and calculating net distance
per_seg = rbind(mCherry.per_seg, MUT2.per_seg, WT.per_seg)
per_seg$net_dist = abs(per_seg$distance)

#calculating mean and sd of per-segment data
# per_seg.mean.df = aggregate(per_seg[,c("velocity", "distance", "net_dist")], by = list(per_seg$cond), FUN = mean)
# per_seg.sd.df = aggregate(per_seg[,c("velocity", "distance", "net_dist")], by = list(per_seg$cond), FUN = sd)
# per_seg.sem.df = aggregate(per_seg[,c("velocity", "distance", "net_dist")], by = list(per_seg$cond), FUN = function(x){sqrt(var(x)/length(x))})

#calculating mean and sd of per-segment data CATEGORISED BY TYPE OF MOVEMENT (i.e. anterograde/retrograde/static)...
# per_seg.cat_mean.df = aggregate(per_seg[,c("velocity", "distance", "net_dist")], by = list(per_seg$cond, per_seg$direction), FUN = mean)
# per_seg.cat_sd.df = aggregate(per_seg[,c("velocity", "distance", "net_dist")], by = list(per_seg$cond, per_seg$direction), FUN = sd)
# per_seg.cat_sem.df = aggregate(per_seg[,c("velocity", "distance", "net_dist")], by = list(per_seg$cond, per_seg$direction), FUN = function(x){sqrt(var(x)/length(x))})
# #...and combining into a single data frame
# per_seg.categorised = full_join(per_seg.cat_mean.df, per_seg.cat_sd.df, by = c("Group.1", "Group.2"))
# per_seg.categorised = full_join(per_seg.categorised, per_seg.cat_sem.df, by = c("Group.1", "Group.2"))
# names(per_seg.categorised) = c("cond", "direction", "velocity_m", "distance_m", "net_dist_m", "velocity_sd", "distance_sd", "net_dist_sd", "velocity_sem", "distance_sem", "net_dist_sem")

# #counting number of categorised movements in each direction and calculating anterograde/retrograde ratio
# counts = summarise(group_by(per_seg, cond, direction),count = n())
# ar_ratios = sapply(unique(counts$cond), function(c){ counts[counts$cond == c & counts$direction == "anterograde", "count"]/counts[counts$cond == c & counts$direction == "retrograde", "count"] })
# 
# #counts skewed by unequal # cells imaged - normalise to number of cells
# #TODO - account for varying length of axon imaged (or at least calculate range)
# counts$norm = unlist(lapply(unique(counts$cond), function(c){ counts[counts$cond == c, "count"]/length(get(paste0(c, "_data"))) }))

#ORDERING
# counts$cond = factor(counts$cond, levels = c("GFP", "WT", "MUT"))
# per_seg.categorised$cond = factor(per_seg.categorised$cond, levels = c("GFP", "WT", "MUT"))
per_seg$cond = factor(per_seg$cond, levels = c("mCherry", "WT", "MUT2"), labels = c("control", "WT", "Active"))
```

```{r warning=FALSE}
per_seg.per_cell = summarise(per_seg %>% group_by(cond, cell, direction, exp), count = n())
#fill in 'missing' values - e.g. count 0 for cells with no anterograde traces
#TODO - need to fill in experiment as well
per_seg.per_cell$cellID = paste0(per_seg.per_cell$exp, per_seg.per_cell$cond, per_seg.per_cell$cell)

directions = c('anterograde', 'retrograde', 'static')
missing_conditions = do.call(rbind, lapply(unique(per_seg.per_cell$cellID), function(c){
  df = per_seg.per_cell[per_seg.per_cell$cellID == c,]
  missing = setdiff(directions, df$direction)
  to_add = data.frame(cond = rep(df$cond[1], length(missing)), 
                      cell = rep(df$cell[1], length(missing)),
                      direction = missing,
                      exp = rep(df$exp[1], length(missing)),
                      count = rep(0, length(missing)),
                      cellID = rep(c, length(missing))
  )
}))

per_seg.per_cell = bind_rows(per_seg.per_cell, missing_conditions)

#ordering
#per_seg.per_cell$cond = factor(per_seg.per_cell$cond, levels = c("GFP", "WT", "MUT"), labels = c("GFP", "WT", "Active")) #already done on per_seg
```

Look at overall number of traces per video - might be skewed by incomplete traces, but large variations should be legitimate (?)
```{r}
mCherry_traces = unlist(sapply(mCherry_Rab7_data, function(d){ d$num_traces }))
Mut2_traces = unlist(sapply(MUT2_Rab7_data, function(d){ d$num_traces }))
WT_traces = unlist(sapply(WT_Rab7_data, function(d){ d$num_traces }))

rab7.traces.df = data.frame(cond = c(rep("mCherry", length(mCherry_traces)), rep("MUT", length(Mut2_traces)), rep("WT", length(WT_traces))),
                       traces = c(mCherry_traces, Mut2_traces, WT_traces), exp = c(unlist(sapply(mCherry_Rab7_data, function(d){ d$experiment })),
                                                                               unlist(sapply(WT_Rab7_data, function(d){ d$experiment })),
                                                                               unlist(sapply(MUT2_Rab7_data, function(d){ d$experiment }))))


#ordering
rab7.traces.df$cond = factor(rab7.traces.df$cond, levels = c("mCherry", "WT", "MUT"), labels = c("control", "WT", "Active"))

```


# Plots

### Average number of particle movements per cell imaged

Dotplot
```{r Rab7_total_movement}
#TODO - fix labels
mov_dot = ggplot(per_seg.per_cell[per_seg.per_cell$direction != "static",], aes(x = cond, y = count, fill = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1)  + 
  geom_boxplot(alpha = 0, width = 0.5, position = position_dodge(0.9), show.legend = F) + 
  labs(x = "", y = "Rab7 movements per cell") + 
  d_format + 
  report_theme + theme(legend.position = 'none') + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell[per_seg.per_cell$direction != "static",]$count), 290, ceiling), size = 3.5)

plot(mov_dot)


#checking p value
d = per_seg.per_cell[per_seg.per_cell$direction == 'anterograde',] %>% group_by(direction)
kruskal.test(count ~ cond, data = d)
```
### LE traces plot
```{r Rab7_trace_dot}
Rab7_trace_dot = ggplot(rab7.traces.df, aes(x = cond, y = traces, fill = cond)) + 
  geom_dotplot(binaxis = 'y', stackdir = 'center', color = NA) + 
  geom_boxplot(alpha = 0, position = position_dodge(width = 0.9), width = 0.2, show.legend = FALSE) + 
  labs(x = "", y = "Total number of LEs", fill = "") + 
  d_format + report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(rab7.traces.df$traces), 10, ceiling), size = 3.5)

plot(Rab7_trace_dot)
```

### 'Run length' plot

Run length violin
```{r run_length_violin, eval = FALSE, echo = FALSE}
run_length_violin = ggplot(per_seg[per_seg$direction!= 'static',], aes(x = cond, y = net_dist, fill = cond, color = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_violin(alpha = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  #d_format + 
  report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg[per_seg$direction!= 'static',]$net_dist), 50, ceiling))

plot(run_length_violin)
```

### Velocity


Velocity violin
```{r velocity_violin, eval = FALSE, echo = FALSE}

velocity_violin = ggplot(per_seg[(abs(per_seg$velocity) < 1000) & per_seg$direction!= "static",], aes(x = cond, fill = cond, y = abs(velocity), colour = cond)) + 
  facet_wrap(~direction) + 
  geom_violin() + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) +
  #ylim(0, 275) + #automatic?
  #d_format + 
  report_theme + theme(legend.position = "none") + 
  stat_compare_means(label = 'p.signif', comparisons = list(c("control", "WT"), c("control", "MUT2")))

plot(velocity_violin)

```


## Per segment analysis --> comparing means per cell

```{r}
per_seg.per_cell.means = do.call(rbind, lapply(1:nrow(per_seg.per_cell), function(i){
  row = per_seg.per_cell[i,];
  all = per_seg[per_seg$cell == row$cell & per_seg$cond == row$cond & per_seg$direction == row$direction,];
  row$mean_dist = mean(all$net_dist);
  row$mean_speed = mean(abs(all$velocity));
  return(row);
}))

#Remove rows with NaN means - 0 cells for condition/direction
per_seg.per_cell.means = per_seg.per_cell.means[!is.nan(per_seg.per_cell.means$mean_dist) & !is.nan(per_seg.per_cell.means$mean_speed),]
```

Distance
```{r Rab7_mean_run_length}
mean_run_length = ggplot(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',], aes(x = cond, y = mean_dist, fill = cond, color = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  d_format + 
  report_theme + theme(legend.position = "none") + 
  labs(x = "", y = "Mean run length") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_dist), 1, ceiling), size = 3.5)

plot(mean_run_length)


```


Velocity means
```{r Rab7_mean_velocity}
velocity_box = ggplot(per_seg.per_cell.means[per_seg.per_cell.means$direction!= "static",], aes(x = cond, fill = cond, y = mean_speed, colour = cond)) + 
  facet_wrap(~direction) + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) +
  labs(x = "", y = "Mean velocity") +
  #ylim(0, 275) + #automatic?
  d_format + 
  report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell.means[per_seg.per_cell.means$direction != "static",]$mean_speed), 1, ceiling), size = 3.5)

plot(velocity_box)

```




COMBINED PLOT
```{r Rab7_full_figure, fig.width = 16, fig.height = 4}
mov_dot + mean_run_length + velocity_box + 
  plot_annotation(tag_levels = 'A') & 
  theme_bw() & 
  theme(legend.position = 'none', 
        panel.background = element_rect(fil = "#FDFDFD"), #custom bg for slide format
        plot.background = element_rect(fil = "#FDFDFD", colour = "#FDFDFD"), 
        legend.background = element_rect(fil = "#FDFDFD"),
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 14, margin = margin(r = 5)),
        strip.text = element_text(size = 12),
        plot.tag = element_text(size = 18, face = 'bold'))
```




# FULL FIGURE

Import representative kymographs
```{r}
#TODO - crop to consistent widths for plot
#normalise measurements to axon length?

mCherry_Rab7_example = read.bitmap("C:\\Users\\fl299\\PhD\\Presentations\\Third\ year\ seminar\\Ptdn_Rab7_representative_kymographs\\CNC61_mCherry_C1.png")
WT_Rab7_example = read.bitmap("C:\\Users\\fl299\\PhD\\Presentations\\Third\ year\ seminar\\Ptdn_Rab7_representative_kymographs\\CNC61_WT_C2.png")
Mut_Rab7_example = read.bitmap("C:\\Users\\fl299\\PhD\\Presentations\\Third\ year\ seminar\\Ptdn_Rab7_representative_kymographs\\CNC61_MUT_C6.png")

mCherry_example.gg = image_as_ggplot(mCherry_Rab7_example) + labs(title = "    Control")
WT_example.gg = image_as_ggplot(WT_Rab7_example) + labs(title = "    WT Protrudin")
Mut_example.gg = image_as_ggplot(Mut_Rab7_example) + labs(title = "    Active Protrudin")
```

Patchwork
```{r Rab7_full_kymographs, fig.width = 16, fig.height = 8}
  (mCherry_example.gg + WT_example.gg + Mut_example.gg)/
  (Rab7_trace_dot + mean_run_length + velocity_box) + 
  plot_annotation(tag_levels = 'A') & 
  report_theme & 
  theme(legend.position = 'none', 
        # panel.background = element_rect(fill = "#FDFDFD", colour = "#FDFDFD"), #custom bg for slide format
        # plot.background = element_rect(fill = "#FDFDFD", colour = "#FDFDFD"), 
        # legend.background = element_rect(fill = "#FDFDFD"),
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 16, margin = margin(r = 5)),
        strip.text = element_text(size = 16),
        plot.tag = element_text(size = 18, face = 'bold'))

```



#Edited for Richard's Karolinska seminar
```{r Karolinska_Rab7, fig.width = 16, fig.height = 9}

(mCherry_example.gg + WT_example.gg + Mut_example.gg + plot_layout(widths = c(ncol(mCherry_Rab7_example)/nrow(mCherry_example),
                                                                                     ncol(WT_Rab7_example)/nrow(WT_example),
                                                                                     ncol(Mut_Rab7_example)/nrow(Mut_example))))/
  (Rab7_trace_dot + mean_run_length + velocity_box) & 
  theme(legend.position = 'none', 
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 16, margin = margin(r = 5)),
        strip.text = element_text(size = 16),
        panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
        plot.title = element_text(size = 18),
        plot.tag = element_text(size = 18, face = 'bold'),
        plot.margin = unit(c(5.5, 5.5, 5.5, 5.5), "points"))


  # (GFP_LAMP1_example.gg + WT_LAMP1_example.gg + Mut_LAMP1_example.gg)/
  # (LAMP1_mov_dot + LAMP1_mean_run_length + LAMP1_mean_run_velocity) &
  # theme(legend.position = 'none', 
  #       axis.text.x = element_text(size = 14),
  #       axis.title.y = element_text(size = 16, margin = margin(r = 5)),
  #       strip.text = element_text(size = 16),
  #       plot.tag = element_text(size = 18, face = 'bold'))


```