# Setup
```{r LAMP1_setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dev = 'png')
knitr::opts_chunk$set(fig.path = '../figures/')


source('C:\\Users\\fl299\\GitHub\\kymograph-analysis\\Kymograph_trace_analysis.R')
source('../style/thesis_setup.R')

library(dplyr)
library(ggplot2)
library(gridExtra)
#library(plyr) #calling explicitly to avoid conflict with dplyr
library(ggpubr)
library(reshape2)
library(PMCMRplus)
library(rstatix)

library(readbitmap)
library(patchwork)
```

# Data handling
```{r LAMP1_csv_selection}

CNC20 = "C:\\Users\\fl299\\GitHub\\first-year-report\\data_analysis\\data\\CNC20_kymo_analysis"
CNC25 = "C:\\Users\\fl299\\GitHub\\first-year-report\\data_analysis\\data\\CNC25_kymo_analysis"
CNC26 = "C:\\Users\\fl299\\GitHub\\first-year-report\\data_analysis\\data\\CNC26_kymo_analysis"
CNC32 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\kymograph_processing_TO_SORT\\CNC32_kymograph_analysis"

#keeping experiments separate because of frame rate differences
GFP_CSVs_20 = list.files(CNC20, pattern = "^GFP_.+\\.csv$", full.names = TRUE)
Mut2_CSVs_20 = list.files(CNC20, pattern = "^M(UT)?2_.+\\.csv$", full.names = TRUE)
WT_CSVs_20 = list.files(CNC20, pattern = "^WT_.+\\.csv$", full.names = TRUE)

GFP_CSVs_25 = list.files(CNC25, pattern = "^GFP_.+\\.csv$", full.names = TRUE)
Mut2_CSVs_25 = list.files(CNC25, pattern = "^M(UT)?2_.+\\.csv$", full.names = TRUE)
WT_CSVs_25 = list.files(CNC25, pattern = "^WT_.+\\.csv$", full.names = TRUE)

GFP_CSVs_26 = list.files(CNC26, pattern = "^GFP_.+\\.csv$", full.names = TRUE)
Mut2_CSVs_26 = list.files(CNC26, pattern = "^M(UT)?2_.+\\.csv$", full.names = TRUE)
WT_CSVs_26 = list.files(CNC26, pattern = "^WT_.+\\.csv$", full.names = TRUE)

GFP_CSVs_32 = list.files(CNC32, pattern = "^GFP_.+\\.csv$", full.names = TRUE)
Mut2_CSVs_32 = list.files(CNC32, pattern = "^M(UT)?2_.+\\.csv$", full.names = TRUE)
WT_CSVs_32 = list.files(CNC32, pattern = "^WT_.+\\.csv$", full.names = TRUE)

#TODO - expand to look at repeat consistency
```


Parse CSVs
```{r analyse_LAMP1_traces}
GFP_data_20 = lapply(GFP_CSVs_20, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 1.1)
  traces$experiment = "CNC20"
  return(traces)
})
Mut2_data_20 = lapply(Mut2_CSVs_20, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 1.1)
  traces$experiment = "CNC20"
  return(traces)
})
WT_data_20 = lapply(WT_CSVs_20, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 1.1)
  traces$experiment = "CNC20"
  return(traces)
})

GFP_data_25 = lapply(GFP_CSVs_25, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC25"
  return(traces)
})
Mut2_data_25 = lapply(Mut2_CSVs_25, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC25"
  return(traces)
})
WT_data_25 = lapply(WT_CSVs_25, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC25"
  return(traces)
})

GFP_data_26 = lapply(GFP_CSVs_26, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC26"
  return(traces)
})
Mut2_data_26 = lapply(Mut2_CSVs_26, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC26"
  return(traces)
})
WT_data_26 = lapply(WT_CSVs_26, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC26"
  return(traces)
})

GFP_data_32 = lapply(GFP_CSVs_32, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC32"
  return(traces)
})
Mut2_data_32 = lapply(Mut2_CSVs_32, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC32"
  return(traces)
})
WT_data_32 = lapply(WT_CSVs_32, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.53)
  traces$experiment = "CNC32"
  return(traces)
})

#combine
GFP_data = c(GFP_data_20, GFP_data_25, GFP_data_26, GFP_data_32)
Mut2_data = c(Mut2_data_20, Mut2_data_25, Mut2_data_26, Mut2_data_32)
WT_data = c(WT_data_20, WT_data_25, WT_data_26, WT_data_32)

```

Basic analysis
```{r}
#all one video per cell in this case - don't need separate count and can number sequentially
n_cells = list("GFP" = length(GFP_data), "WT" = length(WT_data), "MUT" = length(Mut2_data))

#pulling out 'per segment' data from analysed CSVs
GFP.per_seg = do.call(rbind, lapply(1:length(GFP_data), function(c){ df = GFP_data[[c]]$per_seg; df$cell = c; df$cond = "GFP"; df$exp = GFP_data[[c]]$experiment; return(df) }))
Mut2.per_seg = do.call(rbind, lapply(1:length(Mut2_data), function(c){ df = Mut2_data[[c]]$per_seg; df$cell = c; df$cond = "MUT"; df$exp = Mut2_data[[c]]$experiment; return(df) }))
WT.per_seg = do.call(rbind, lapply(1:length(WT_data), function(c){ df = WT_data[[c]]$per_seg; df$cell = c; df$cond = "WT"; df$exp = WT_data[[c]]$experiment; return(df) }))

#combining data from all conditions and calculating net distance
per_seg = rbind(GFP.per_seg, Mut2.per_seg, WT.per_seg)
per_seg$net_dist = abs(per_seg$distance)


#ORDERING
per_seg$cond = factor(per_seg$cond, levels = c("GFP", "WT", "MUT"), labels = c("Control", "WT", "Active"))
```

```{r warning=FALSE}
per_seg.per_cell = summarise(per_seg %>% group_by(cond, cell, direction, exp), count = n())
#fill in 'missing' values - e.g. count 0 for cells with no anterograde traces
#TODO - need to fill in experiment as well
per_seg.per_cell$cellID = paste0(per_seg.per_cell$exp, per_seg.per_cell$cond, per_seg.per_cell$cell)

directions = c('anterograde', 'retrograde', 'static')
missing_conditions = do.call(rbind, lapply(unique(per_seg.per_cell$cellID), function(c){
  df = per_seg.per_cell[per_seg.per_cell$cellID == c,]
  missing = setdiff(directions, df$direction)
  to_add = data.frame(cond = rep(df$cond[1], length(missing)), 
                      cell = rep(df$cell[1], length(missing)),
                      direction = missing,
                      exp = rep(df$exp[1], length(missing)),
                      count = rep(0, length(missing)),
                      cellID = rep(c, length(missing))
  )
}))

per_seg.per_cell = bind_rows(per_seg.per_cell, missing_conditions)

```

# Plots

### Average number of particle movements per cell imaged

Dotplot
```{r LAMP1_mov_dot}
#TODO - fix labels
LAMP1_mov_dot = ggplot(per_seg.per_cell[per_seg.per_cell$direction != "static",], aes(x = cond, y = count, fill = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1)  + 
  geom_boxplot(alpha = 0, width = 0.5, position = position_dodge(0.9), show.legend = F) + 
  labs(x = "", y = "LAMP1 movements per cell") + 
  d_format + 
  report_theme + theme(legend.position = 'none') + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell[per_seg.per_cell$direction != "static",]$count), 25, ceiling), size = p_fs)

plot(LAMP1_mov_dot)
```
analysis validation
```{r}
car::leveneTest(count~cond, data = per_seg.per_cell[per_seg.per_cell$direction == 'anterograde',])
car::leveneTest(count~cond, data = per_seg.per_cell[per_seg.per_cell$direction == 'retrograde',])
#both p > 0.05; no significant variance

antero.aov = aov(count~cond, data = per_seg.per_cell[per_seg.per_cell$direction == 'anterograde',])
retro.aov = aov(count~cond, data = per_seg.per_cell[per_seg.per_cell$direction == 'retrograde',])

summary(antero.aov)
summary(retro.aov)

shapiro.test(per_seg.per_cell[per_seg.per_cell$direction == 'anterograde',]$count)
shapiro.test(per_seg.per_cell[per_seg.per_cell$direction == 'retrograde',]$count)
#both p < 0.05 - NOT NORMAL - Kruskal-Wallis instead of ANOVA
```



### 'Run length' plot

Run length violin
```{r LAMP1_run_length_violin}
run_length_violin = ggplot(per_seg[per_seg$direction!= 'static',], aes(x = cond, y = net_dist, fill = cond, color = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_violin(alpha = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  d_format + report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg[per_seg$direction!= 'static',]$net_dist), 50, ceiling))

plot(run_length_violin)
```
analysis valiadtion - CHECK WITH STATS CLINIC
```{r}
#too far off normal for levene test
fligner.test(net_dist~cond, data = per_seg[per_seg$direction == 'anterograde',])
#p < 0.05
fligner.test(net_dist~cond, data = per_seg[per_seg$direction == 'retrograde',])
#p > 0.05

# antero.aov = aov(net_dist~cond, data = per_seg[per_seg$direction == 'anterograde',])
# retro.aov = aov(net_dist~cond, data = per_seg[per_seg$direction == 'retrograde',])
# 
# summary(antero.aov)
# summary(retro.aov)
# 
shapiro.test(per_seg[per_seg$direction == 'anterograde',]$net_dist)
#p < 0.05 - NOT NORMAL
# shapiro.test(per_seg[per_seg$direction == 'retrograde',]$net_dist)#TOO LARGE FOR SHAPIRO TEST, but pretty clearly not normal
#Kruskal-Wallis instead of ANOVA
```



Look at overall number of traces per video - might be skewed by incomplete traces, but large variations should be legitimate.
```{r}
GFP_traces = unlist(sapply(GFP_data, function(d){ d$num_traces }))
Mut2_traces = unlist(sapply(Mut2_data, function(d){ d$num_traces }))
WT_traces = unlist(sapply(WT_data, function(d){ d$num_traces }))

traces.df = data.frame(cond = c(rep("GFP", length(GFP_traces)), rep("MUT", length(Mut2_traces)), rep("WT", length(WT_traces))),
                       traces = c(GFP_traces, Mut2_traces, WT_traces), exp = c(unlist(sapply(GFP_data, function(d){ d$experiment })),
                                                                               unlist(sapply(WT_data, function(d){ d$experiment })),
                                                                               unlist(sapply(Mut2_data, function(d){ d$experiment }))))

# traces.sum.df = data.frame(cond = c("GFP", "Mut2", "WT"),
#                            mean_traces = c(mean(GFP_traces), mean(Mut2_traces), mean(WT_traces)),
#                            sd_traces = c(sd(GFP_traces), sd(Mut2_traces), sd(WT_traces)))

#ordering
traces.df$cond = factor(traces.df$cond, levels = c("GFP", "WT", "MUT"), labels = c("Control", "WT", "Active"))

```

### Plots of trace #

Dotplot
```{r LAMP1_trace_dot}
LAMP1_trace_dot = ggplot(traces.df, aes(x = cond, y = traces, fill = cond)) + 
  geom_dotplot(binaxis = 'y', stackdir = 'center', color = NA) + 
  geom_boxplot(alpha = 0, position = position_dodge(width = 0.9), width = 0.2, show.legend = FALSE) + 
  labs(x = "", y = "Total number of lysosomes", fill = "") + 
  d_format + report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(traces.df$traces), 10, ceiling), size = p_fs)

plot(LAMP1_trace_dot)
```
analysis validation
```{r}
car::leveneTest(traces~cond, data = traces.df)
car::leveneTest(traces~cond, data = traces.df)
#both p > 0.05; no significant variance


shapiro.test(traces.df$traces)
shapiro.test(traces.df$traces)
#both p < 0.05 - NOT NORMAL - Kruskal-Wallis instead of ANOVA

```



### Velocity


Velocity violin
```{r LAMP1_velocity_violin}

velocity_violin = ggplot(per_seg[(abs(per_seg$velocity) < 1000) & per_seg$direction!= "static",], aes(x = cond, fill = cond, y = abs(velocity), colour = cond)) + 
  facet_wrap(~direction) + 
  geom_violin() + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) +
  #ylim(0, 275) + #automatic?
  d_format + report_theme + theme(legend.position = "none") + 
  stat_compare_means(label = 'p.signif', comparisons = list(c("Control", "WT"), c("Control", "Active")))

plot(velocity_violin)

```

analysis
```{r}
#too far off normal for levene test
fligner.test(abs(velocity)~cond, data = per_seg[(abs(per_seg$velocity) < 1000) & per_seg$direction!= "static",])
fligner.test(abs(velocity)~cond, data = per_seg[(abs(per_seg$velocity) < 1000) & per_seg$direction!= "static",])
#p < 0.05 both - significantly different variance - NEED TO LOOK @ MEAN PER CELL

```


## Per segment analysis --> comparing means per cell

```{r}
per_seg.per_cell.means = do.call(rbind, lapply(1:nrow(per_seg.per_cell), function(i){
  row = per_seg.per_cell[i,];
  all = per_seg[per_seg$cell == row$cell & per_seg$cond == row$cond & per_seg$direction == row$direction,];
  row$mean_dist = mean(all$net_dist);
  row$mean_speed = mean(abs(all$velocity));
  return(row);
}))

#Remove rows with NaN means - 0 cells for condition/direction
per_seg.per_cell.means = per_seg.per_cell.means[!is.nan(per_seg.per_cell.means$mean_dist) & !is.nan(per_seg.per_cell.means$mean_speed),]
```

Distance
```{r LAMP1_mean_run_length}
LAMP1_mean_run_length = ggplot(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',], aes(x = cond, y = mean_dist, fill = cond, color = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(binaxis = 'y', stackdir = 'center', color = NA) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  d_format + report_theme + theme(legend.position = "none") + 
  labs(x = "", y = "Mean run length") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_dist), 30, ceiling), size = p_fs)

plot(LAMP1_mean_run_length)


```
analysis validation
```{r}
car::leveneTest(mean_dist~cond, data = per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',])
car::leveneTest(mean_dist~cond, data = per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',])
#both p > 0.05; no significant variance


shapiro.test(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_dist)
shapiro.test(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_dist)
#both p < 0.05 - NOT NORMAL - Kruskal-Wallis instead of ANOVA
```


Velocity

```{r LAMP1_mean_run_velocity}
#first pass showing K-E p < 0.05 for anterograde only - getting p values for Dunn's test to plot
# ant = data.frame(cond = per_seg.per_cell.means[per_seg.per_cell.means$direction == 'anterograde',]$cond,
#                  mean_speed = per_seg.per_cell.means[per_seg.per_cell.means$direction == 'anterograde',]$mean_speed,
#                  direction = per_seg.per_cell.means[per_seg.per_cell.means$direction == 'anterograde',]$direction)#shouldn't be necessary, but fixes column not found error in dunn_test

ant = per_seg.per_cell.means[per_seg.per_cell.means$direction== 'anterograde',]
test <- ant %>% group_by(direction) %>% dplyr::select(mean_speed, cond)#needs dplyr to work - clash with another package?
LAMP1_mean_run_velocity.dunn = dunn_test(test, mean_speed ~ cond, p.adjust.method = 'bonferroni')
LAMP1_mean_run_velocity.dunn = add_xy_position(LAMP1_mean_run_velocity.dunn, x = "cond")
LAMP1_mean_run_velocity.dunn$p.adj.rounded = sapply(LAMP1_mean_run_velocity.dunn$p.adj, function(p){round(p, digits = 3)})


LAMP1_mean_run_velocity = ggplot(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',], aes(x = cond, y = mean_speed)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(aes(fill = cond, color = cond), binaxis = 'y', stackdir = 'center') + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  d_format + report_theme + 
  labs(x = "", y = "Mean velocity") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_speed), 12, ceiling), size = p_fs) +
  stat_pvalue_manual(LAMP1_mean_run_velocity.dunn, hide.ns = TRUE, label = 'p.adj.rounded', step.increase = 0.1)

plot(LAMP1_mean_run_velocity)

```

analysis valiadtion
```{r}
car::leveneTest(mean_speed~cond, data = per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',])
car::leveneTest(mean_speed~cond, data = per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',])
#both p > 0.05; no significant variance


shapiro.test(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_speed)
shapiro.test(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_speed)
#both p < 0.05 - NOT NORMAL - Kruskal-Wallis instead of ANOVA
```


## Combined figure
```{r LAMP1_full_plots, fig.width = 16, fig.height = 4}
LAMP1_mov_dot + #trace_dot 
LAMP1_mean_run_length + LAMP1_mean_run_velocity + 
  plot_annotation(tag_levels = 'A') & 
  theme_bw() & 
  theme(legend.position = 'none', 
        panel.background = element_rect(fil = "#FDFDFD"), #custom bg for slide format
        plot.background = element_rect(fil = "#FDFDFD", colour = "#FDFDFD"), 
        legend.background = element_rect(fil = "#FDFDFD"),
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 14, margin = margin(r = 5)),
        strip.text = element_text(size = 12),
        plot.tag = element_text(size = 18, face = 'bold'))
```

#---------------------------------------------------------------
#-----------OVERALL VALIDATION - DIFFERENCES BY CNC #-----------
#---------------------------------------------------------------


```{r}
mov.df = per_seg.per_cell[,c('cond', 'direction', 'exp', 'count')] %>% dplyr::group_by(cond, exp, direction) %>% dplyr::summarise(mov = mean(count))
tr.df = traces.df %>% dplyr::group_by(cond, exp) %>% dplyr::summarise(tr = mean(traces))
run_vel.df = per_seg.per_cell.means[,c('cond', 'direction', 'exp', 'mean_dist', 'mean_speed')]%>% dplyr::group_by(cond, exp, direction) %>% dplyr::summarise(run = mean(mean_dist), vel = mean(mean_speed))
```

```{r LAMP1_fridman_mov}

mov.mat.antero = acast(mov.df[mov.df$direction == "anterograde", ], formula = cond ~ exp)
mov.mat.retro = acast(mov.df[mov.df$direction == "retrograde", ], formula = cond ~ exp)
mov.mat.static = acast(mov.df[mov.df$direction == "static", ], formula = cond ~ exp)

friedman.test(mov.mat.antero)
friedman.test(mov.mat.retro)
friedman.test(mov.mat.static)

#none significant
```

```{r LAMP1_friedman_tr}
tr.mat = acast(tr.df, formula = cond ~ exp)

friedman.test(tr.mat)
```

```{r LAMP1_friedman_run}
run.mat.antero = acast(run_vel.df[run_vel.df$direction == 'anterograde',], formula = cond ~ exp, value.var = 'run')
run.mat.retro = acast(run_vel.df[run_vel.df$direction == 'retrograde',], formula = cond ~ exp, value.var = 'run')
run.mat.static = acast(run_vel.df[run_vel.df$direction == 'static',], formula = cond ~ exp, value.var = 'run')

friedman.test(run.mat.antero)
friedman.test(run.mat.retro)
friedman.test(run.mat.static)

#run.mat.retro p = 0.02929
frdAllPairsNemenyiTest(run.mat.retro) #CNC20/CNC26 p = 0.023 #TODO - how to handle? exclude CNC20?
```

```{r LAMP1_friedman_vel}
vel.mat.antero = acast(run_vel.df[run_vel.df$direction == 'anterograde',], formula = cond ~ exp, value.var = 'vel')
vel.mat.retro = acast(run_vel.df[run_vel.df$direction == 'retrograde',], formula = cond ~ exp, value.var = 'vel')
vel.mat.static = acast(run_vel.df[run_vel.df$direction == 'static',], formula = cond ~ exp, value.var = 'vel')

friedman.test(vel.mat.antero)
friedman.test(vel.mat.retro)
friedman.test(vel.mat.static)

#vel.mat.static p = 0.04205
frdAllPairsNemenyiTest(vel.mat.static) #none significant, CNC20/CNC26 p = 0.055
```

# Repetition of analysis excluding CNC20

Quickly checked by commenting out, mean speed for retrograde transport becomes significant (K.W. p = 0.0071), all others n.s.
GFP-WT - p = 0.048
GFP-MUT - p = 0.0013

Not sure if enough evidence to exclude CNC20
Check at next stats clinic?

Repeat with mCherry Ptdn constructs and LAMP1-mNeonGreen to compare












# FULL FIGURE

Import representative kymographs
```{r}
LAMP1_kymo_panel = image_as_ggplot(read.bitmap('C:\\Users\\fl299\\PhD\\LAMP1_kymo_panel.png'))
```

Patchwork
```{r LAMP1_full_figure_def, fig.width=16, fig.height=12}
  LAMP1_full = LAMP1_kymo_panel/
  (LAMP1_trace_dot + LAMP1_mean_run_length + LAMP1_mean_run_velocity) + 
  plot_annotation(tag_levels = 'A')

LAMP1_full
```

