---
output:
  pdf_document: default
  html_document: default
---
# Setup
```{r setup}
knitr::opts_chunk$set(echo = TRUE, dev = 'png')
knitr::opts_chunk$set(fig.path = "../figures/")


source('../scripts_and_macros/Kymograph_trace_analysis.R')#copied over from git repo 08-02-23
source('../style/thesis_setup.R')


library(dplyr)
library(ggplot2)
library(gridExtra)
#library(plyr) #calling explicitly to avoid conflict with dplyr
library(ggpubr)
library(reshape2)
library(PMCMRplus)
library(here)

library(readbitmap)
library(patchwork)
```

# Data handling
```{r Rab7_csv_selection}

# CNC57 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC57"
# CNC58_59 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC58-59"
# CNC59 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC59"
# CNC61 = "C:\\Users\\fl299\\PhD\\Data\\Mine\\Spinning Disc\\Kymographs and Traces\\traced\\CNC61"

CNC57 = "../data/kymograph_traces/CNC57/"
CNC58_59 = "../data/kymograph_traces/CNC58-59/"
CNC59 = "../data/kymograph_traces/CNC59/"
CNC61 = "../data/kymograph_traces/CNC61/"

#keeping experiments separate because of frame rate differences


mCherry_CSVs_58 = list.files(CNC58_59, pattern = "^MCR_.+\\.csv$", full.names = TRUE)
WT_CSVs_58 = list.files(CNC58_59, pattern = "^WTR_.+\\.csv$", full.names = TRUE)
MUT2_CSVs_58 = list.files(CNC58_59, pattern = "^MR_.+\\.csv$", full.names = TRUE)


mCherry_CSVs_59 = list.files(CNC59, pattern = "^MCR_.+\\.csv$", full.names = TRUE)
WT_CSVs_59 = list.files(CNC59, pattern = "^WTR_.+\\.csv$", full.names = TRUE)
MUT2_CSVs_59 = list.files(CNC59, pattern = "^MR_.+\\.csv$", full.names = TRUE)

mCherry_CSVs_61 = list.files(CNC61, pattern = "^MCR_.+\\.csv$", full.names = TRUE)
WT_CSVs_61 = list.files(CNC61, pattern = "^WTR_.+\\.csv$", full.names = TRUE)
MUT2_CSVs_61 = list.files(CNC61, pattern = "^M2R_.+\\.csv$", full.names = TRUE)


```


```{r analyse_Rab7_traces}
#TODO - fix frame rates to CSV for each stack


mCherry_data_58 = lapply(mCherry_CSVs_58, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC58"
  return(traces)
})
WT_data_58 = lapply(WT_CSVs_58, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC58"
  return(traces)
})
MUT2_data_58 = lapply(MUT2_CSVs_58, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC58"
  return(traces)
})

mCherry_data_59 = lapply(mCherry_CSVs_59, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC59"
  return(traces)
})
WT_data_59 = lapply(WT_CSVs_59, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC59"
  return(traces)
})
MUT2_data_59 = lapply(MUT2_CSVs_59, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC59"
  return(traces)
})

mCherry_data_61 = lapply(mCherry_CSVs_61, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC61"
  return(traces)
})
WT_data_61 = lapply(WT_CSVs_61, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC61"
  return(traces)
})
MUT2_data_61 = lapply(MUT2_CSVs_61, function(c){
  traces = analyse_traces(csv_path = c, x_scale = 0.16, y_scale = 0.56, mov_threshold = 5)
  traces$experiment = "CNC61"
  return(traces)
})





#combine


mCherry_Rab7_data = c(mCherry_data_58, mCherry_data_59, mCherry_data_61)
WT_Rab7_data = c(WT_data_58, WT_data_59, WT_data_61)
MUT2_Rab7_data = c(MUT2_data_58, MUT2_data_59, MUT2_data_61)



```


Basic analysis
```{r}
#all one video per cell in this case - don't need separate count and can number sequentially - LEAVING ARL8 FOR NOW
n_cells = list(mCherry = length(mCherry_Rab7_data), WT = length(WT_Rab7_data), MUT2 = length(MUT2_Rab7_data))

#pulling out 'per segment' data from analysed CSVs
mCherry.per_seg = do.call(rbind, lapply(1:length(mCherry_Rab7_data), function(c){ df = mCherry_Rab7_data[[c]]$per_seg; df$cell = c; df$cond = "mCherry"; df$exp = mCherry_Rab7_data[[c]]$experiment; return(df) }))
WT.per_seg = do.call(rbind, lapply(1:length(WT_Rab7_data), function(c){ df = WT_Rab7_data[[c]]$per_seg; df$cell = c; df$cond = "WT"; df$exp = WT_Rab7_data[[c]]$experiment; return(df) }))
MUT2.per_seg = do.call(rbind, lapply(1:length(MUT2_Rab7_data), function(c){ df = MUT2_Rab7_data[[c]]$per_seg; df$cell = c; df$cond = "MUT2"; df$exp = MUT2_Rab7_data[[c]]$experiment; return(df) }))


#combining data from all conditions and calculating net distance
per_seg = rbind(mCherry.per_seg, MUT2.per_seg, WT.per_seg)
per_seg$net_dist = abs(per_seg$distance)


#ORDERING
per_seg$cond = factor(per_seg$cond, levels = c("mCherry", "WT", "MUT2"), labels = c("Control", "WT", "Active"))
```

```{r warning=FALSE}
per_seg.per_cell = summarise(per_seg %>% group_by(cond, cell, direction, exp), count = n())
#fill in 'missing' values - e.g. count 0 for cells with no anterograde traces
#TODO - need to fill in experiment as well
per_seg.per_cell$cellID = paste0(per_seg.per_cell$exp, per_seg.per_cell$cond, per_seg.per_cell$cell)

directions = c('anterograde', 'retrograde', 'static')
missing_conditions = do.call(rbind, lapply(unique(per_seg.per_cell$cellID), function(c){
  df = per_seg.per_cell[per_seg.per_cell$cellID == c,]
  missing = setdiff(directions, df$direction)
  to_add = data.frame(cond = rep(df$cond[1], length(missing)), 
                      cell = rep(df$cell[1], length(missing)),
                      direction = missing,
                      exp = rep(df$exp[1], length(missing)),
                      count = rep(0, length(missing)),
                      cellID = rep(c, length(missing))
  )
}))

per_seg.per_cell = bind_rows(per_seg.per_cell, missing_conditions)
```

Look at overall number of traces per video - might be skewed by incomplete traces, but large variations should be legitimate (?)
```{r}
mCherry_traces = unlist(sapply(mCherry_Rab7_data, function(d){ d$num_traces }))
Mut2_traces = unlist(sapply(MUT2_Rab7_data, function(d){ d$num_traces }))
WT_traces = unlist(sapply(WT_Rab7_data, function(d){ d$num_traces }))

rab7.traces.df = data.frame(cond = c(rep("mCherry", length(mCherry_traces)), rep("MUT", length(Mut2_traces)), rep("WT", length(WT_traces))),
                       traces = c(mCherry_traces, Mut2_traces, WT_traces), exp = c(unlist(sapply(mCherry_Rab7_data, function(d){ d$experiment })),
                                                                               unlist(sapply(WT_Rab7_data, function(d){ d$experiment })),
                                                                               unlist(sapply(MUT2_Rab7_data, function(d){ d$experiment }))))


#ordering
rab7.traces.df$cond = factor(rab7.traces.df$cond, levels = c("mCherry", "WT", "MUT"), labels = c("Control", "WT", "Active"))

```


# Plots

### Average number of particle movements per cell imaged

Dotplot
```{r Rab7_total_movement}
#TODO - fix labels
mov_dot = ggplot(per_seg.per_cell[per_seg.per_cell$direction != "static",], aes(x = cond, y = count, fill = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1)  + 
  geom_boxplot(alpha = 0, width = 0.5, position = position_dodge(0.9), show.legend = F) + 
  labs(x = "", y = "Rab7 movements per cell") + 
  d_format + report_theme +
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell[per_seg.per_cell$direction != "static",]$count), 290, ceiling), size = p_fs)

plot(mov_dot)


#checking p value
d = per_seg.per_cell[per_seg.per_cell$direction == 'anterograde',] %>% group_by(direction)
kruskal.test(count ~ cond, data = d)
```
### LE traces plot
```{r Rab7_trace_dot}
Rab7_trace_dot = ggplot(rab7.traces.df, aes(x = cond, y = traces, fill = cond)) + 
  geom_dotplot(binaxis = 'y', stackdir = 'center', color = NA) + 
  geom_boxplot(alpha = 0, position = position_dodge(width = 0.9), width = 0.2, show.legend = FALSE) + 
  labs(x = "", y = "Total number of LEs", fill = "") + 
  d_format + report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(rab7.traces.df$traces), 350, ceiling), size = p_fs)

plot(Rab7_trace_dot)
```

### 'Run length' plot

Run length violin
```{r Rab7_run_length_violin, eval = FALSE, echo = FALSE}
run_length_violin = ggplot(per_seg[per_seg$direction!= 'static',], aes(x = cond, y = net_dist, fill = cond, color = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_violin(alpha = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  #d_format + 
  report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg[per_seg$direction!= 'static',]$net_dist), 50, ceiling))

plot(run_length_violin)
```

### Velocity


Velocity violin
```{r Rab7_velocity_violin, eval = FALSE, echo = FALSE}

velocity_violin = ggplot(per_seg[(abs(per_seg$velocity) < 1000) & per_seg$direction!= "static",], aes(x = cond, fill = cond, y = abs(velocity), colour = cond)) + 
  facet_wrap(~direction) + 
  geom_violin() + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) +
  #ylim(0, 275) + #automatic?
  #d_format + 
  report_theme + theme(legend.position = "none") + 
  stat_compare_means(label = 'p.signif', comparisons = list(c("Control", "WT"), c("Control", "MUT2")))

plot(velocity_violin)

```


## Per segment analysis --> comparing means per cell

```{r}
per_seg.per_cell.means = do.call(rbind, lapply(1:nrow(per_seg.per_cell), function(i){
  row = per_seg.per_cell[i,];
  all = per_seg[per_seg$cell == row$cell & per_seg$cond == row$cond & per_seg$direction == row$direction,];
  row$mean_dist = mean(all$net_dist);
  row$mean_speed = mean(abs(all$velocity));
  return(row);
}))

#Remove rows with NaN means - 0 cells for condition/direction
per_seg.per_cell.means = per_seg.per_cell.means[!is.nan(per_seg.per_cell.means$mean_dist) & !is.nan(per_seg.per_cell.means$mean_speed),]
```

Distance
```{r Rab7_mean_run_length}
mean_run_length = ggplot(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',], aes(x = cond, y = mean_dist, fill = cond, color = cond)) + 
  facet_wrap(~direction, strip.position = 'top') + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) + 
  d_format + 
  report_theme + theme(legend.position = "none") + 
  labs(x = "", y = paste0("Mean run length (", "\U03BC", "m)")) + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell.means[per_seg.per_cell.means$direction!= 'static',]$mean_dist), 1, ceiling), size = p_fs)

plot(mean_run_length)


```


Velocity means
```{r Rab7_mean_velocity}
velocity_box = ggplot(per_seg.per_cell.means[per_seg.per_cell.means$direction!= "static",], aes(x = cond, fill = cond, y = mean_speed, colour = cond)) + 
  facet_wrap(~direction) + 
  geom_dotplot(binaxis = "y", position = "dodge", stackdir = "center", width = 0.5, color = NA, dotsize = 1) + 
  geom_boxplot(alpha = 0, colour = "black", position = position_dodge(width = 0.9), width = 0.5, show.legend = FALSE) +
  labs(x = "", y = paste0("Mean velocity (", "\U03BC", "m/s)")) +
  #ylim(0, 275) + #automatic?
  d_format + 
  report_theme + theme(legend.position = "none") + 
  stat_compare_means(label.y = plyr::round_any(max(per_seg.per_cell.means[per_seg.per_cell.means$direction != "static",]$mean_speed), 1, ceiling), size = p_fs)

plot(velocity_box)

```




COMBINED PLOT
```{r Rab7_full_plots, fig.width = 16, fig.height = 4}
mov_dot + mean_run_length + velocity_box + 
  plot_annotation(tag_levels = 'A') & 
  theme_bw() & 
  theme(legend.position = 'none', 
        panel.background = element_rect(fil = "#FDFDFD"), #custom bg for slide format
        plot.background = element_rect(fil = "#FDFDFD", colour = "#FDFDFD"), 
        legend.background = element_rect(fil = "#FDFDFD"),
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 14, margin = margin(r = 5)),
        strip.text = element_text(size = 12),
        plot.tag = element_text(size = 18, face = 'bold'))
```




# FULL FIGURE

Import representative kymographs
```{r}
Rab7_kymo_panel = image_as_ggplot(read.bitmap(here('figures/external/Rab7_kymo_panel.png')))
```

Patchwork
```{r Rab7_full_figure_def}
RAB7_full = Rab7_kymo_panel/
  (Rab7_trace_dot + mean_run_length + velocity_box) + 
  plot_annotation(tag_levels = 'A')

RAB7_full
```


