---
output:
  word_document: default
  html_document: default
  pdf_document: 
    latex_engine: xelatex
bibliography: "`r here::here('bib/THESIS.bib')`"
---

# Protrudin's effect on neuroprotection through lysosome and late endosome transport

```{r chapter_1_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

knitr::knit_child(here::here("analysis/ONC_figure.Rmd"))
knitr::knit_child(here::here("analysis/Ptdn_LAMP1_analysis.Rmd"))
knitr::knit_child(here::here("analysis/Ptdn_Rab7_analysis.Rmd"))
knitr::knit_child(here::here("analysis/FYCO1_RNAseq_analysis.Rmd"))


library(knitr)

```

## Introduction

### The many roles of neuronal lysosomes

Lysosomes have traditionally been understood as the degradative compartment of the cell, responsible for the breakdown and recycling of excess or damaged proteins and organelles.  While this is certainly true, more recent work has shown that they are in fact a diverse pool of complex, multi-functional organelles.  While lysosomal function is essential to cellular survival broadly, it is particularly critical in neurons, which have additional metabolic, signalling, and transport constraints imposed by their extreme morphology.  As a result, normal activity and transport of lysosomes is a key factor in neuroprotection, and disruption of either of these processes can lead to neurodegeneration.

Lysosomal dysfunction is implicated in a wide range of neurodegenerative diseases, including Alzheimers, Parkinson's, Huntington's, and ALS.  In many cases, defects in lysosomal degradation of autophagosome contents (a process which occurs in tandem with retrograde transport in axons, as reviewed in [@fergusonNeuronalLysosomes2019]) are thought to cause the accumulation of damaged organelles, which eventually leads to axonal swellings and subseqent degeneration [@dehayPathogenicLysosomalDepletion2010][@leeLysosomalProteolysisInhibition2011][@xieEndolysosomalDeficitsAugment2015].  Other conditions have been linked to impaired transport leading to decreased availability of lysosomes in the axon [@roneyLipidmediatedMotoradaptorSequestration2021].  Understanding how lysosomes function in neurons and the processes controlling their transport is therefore an important step in developing new protective or regenerative therapies for the CNS.


#### RNA transport

In recent years, a substantial body of evidence has emerged for lysosomes and closely related late endosomes facilitating RNA transport in neurons.  mRNA transport, and subsequent local protein synthesis, is essential for axon growth, pathfinding, maintenance, and synapse formation (reviewed in [@cioniMolecularControlLocal2018]). 

Cioni et al. demonstrated that ribonucleoproteins in the axon often associate with Rab7-marked late endosomes, and that these ensosomes act as sites for local translation.  Furthermore, these late endosomes can form contact sites with mitochondria, where they are involved in the local translation of proteins for mitochondrial maintenance [@cioniLateEndosomesAct2019].  Work by Liao et al. identified the protein ANXA11 as the tether between lysosomes and RNA granules in the axon, and also showed that ALS-associated mutation in ANXA11 interfere with lysosome binding [@liaoRNAGranulesHitchhike2019].  


#### Signalling platform

As an endpoint of several protein degradation pathways, lysosomes are particularly responsive to amino acid levels in the cell, and in turn regulate the cell's response to feeding or starvation through signalling complex mTORC1.  mTORC1 localises to lysosomes, where it can be activated by regulatory proteins that colocalise in the presence of amino acids, in turn leading to increased protein translation and cell growth (reviewed in [@inpanathanLysosomeSignalingPlatform2019]).This process is modulated by lysosome positioning - during starvation, lysosomes cluster near the nucleus, and mTORC1 activity is inhibited.  Feeding allows the lysosomes to transport towards the cell periphery, where mTORC1 actiavation can protmote growth.  Hong et al. found that protrudin was involved in this process, interacting with lysosomes to facilitate anterograde transport [@hongPtdIns3PControlsMTORC12017].   This signalling is especially important in neurons - [@poulopoulosSubcellularTranscriptomesProteomes2019] found that the majority of mRNA transcripts that they idenfitied as enriched in growth cones contain a TOP motif for mTORC1-dependent translation.  Therefore, the location of lysosomes and the regulation of mTORC1 on those lysosomes likely acts as a key gatekeeper of local translation, and the associated axonal functions that depend on it.

Beyond mTORC1, lysosomes also participate in calcium signalling, which is particularly important for neuronal function.  Lysosomes are known to act as calcium stores for the cell, but also participate in store-operated calcium entry (SOCE) in response to calcium influx.  In this process, lysosomes are exocytosed, inserting N-type calcium channels into the plasma membrane to allow increased calcium entry and subsequent activation of signalling cascades [@huiReleaseCalciumEndolysosomes2015].  Additionally, this exocytosis releases cathepsin B into the extracellular space, which activates matrix metalloproteinase 9 (MMP-9) through cleaving its inhibitor TIMP-1.  Active MMP-9 breaks down components of the extracellular matrix, facilitating cellular growth.  In neurons, this allows dendritic spines to grow in response to activity [@padamseyActivityDependentExocytosisLysosomes2017].  In cancer cells, this underpins the growth of invasive protrusions called invadopodia, a process that has been shown to depend on protrudin, Rab7, and FYCO1 [@pedersenProtrudinmediatedEREndosome2020]. 


#### ER morphology

Recent advances in microscopy have allowed the interactions between lysosomes and the ER to be imaged in unprecedented detail, and shown how lysosomes are invovled in shaping the tubular ER network in cells.   Lu et al. demonstrated that, lysosomes moved in response to the cell's nutritional state, they pulled ER tubules with them.  Interfering with the anterograde movement of lysosomes or ER-lysosomal contact sites resulted in fewer tubules and large ER sheets near the nucleus.  In neurons, this resulted in decreased axonal ER tubules and stunted axon growth [@luStructureGlobalDistribution2020].  Spits et al. showed a similar interaction with late endosomes, where Rab7 and its transport adaptors mediated ER tubule reorganisation, and resulting in a more sparse ER network in response to starvation [@spitsMobileLateEndosomes2021].

As much as lysosomes help shape the ER, these interactions are reciprocal, and also play an essential role in lysosome morphology.  Disruption of the ER tubule network has been shown to result in the accumulation of large, static lysosomes near the nucleus, and prevent lysosome transport into axons [@ozkanERLysosomeContacts2021].  Other work has shown that association with ER tubules is necessary for lysosomal tubulation and cargo sorting, and that a key protein involved, LRRK2, is implicated in Parkinson's disease [@bonet-ponceEndoplasmicReticulumContributes2022].



### Lysosome transport mechanisms

As lysosomal function is closely tied to the positioning of these organelles in the cell, significant work has gone into understanding how lysosomes are transported.  This has identified a number of distinct, sometimes overlapping pathways that are regulated in different ways.  Work is still ongoing to understand the importance of these different pathways, and how they may interact in different cell types to control late endosome and lysosomal transport.


#### BORC/Arl8b/SKIP

The BORC/Arl8b/SKIP interaction is one of the best-characterised mechanisms involved in lysosome transport, in part because of how many different proteins are involved and their roles in other systems.  In summary, BORC, a complex of eight proteins, is recruited to the lysosomal membrane, where it in turn recruits and activates Arl8b, a small GTPase.  GTP-bound Arl8b can in turn interact with SKIP, which binds to kinesin-1.  This chain of interactions tethers lysosomes to the kinesin motor for transport along microtubules [@guardiaBORCFunctionsUpstream2016].  This process has been shown to mediate anterograde transport of lysosomes in axons [@fariasBORCKinesin1Ensemble2017], and defects in BORC component diaskedin are linked to progressive axonal dystrophy [@snouwaertMutationBorcs7Subunit2018].  In cancer cells, this system has been shown to enhance cell invasiveness, supporting a role for BORC-mediated lysosome transport in cell outgrowth [@wuLysosomalTraffickingMediated2020].


#### Calcyon and other adaptor proteins

While the BORC complex is invovled in kinesin-mediated transport, several other adaptor proteins facilitate minus-end directed transport through binding to dyneins.  These include calcyon (which binds to late endosomes and lysosomes)[@shiDyneinBindsStimulates2017][@shiCouplingMicrotubuleMotors2018], RILP (which is a specific adaptor for Rab)[@bucciExpressionAnalysisChromosomal2001], ALG2 (which binds lysosomes in a calcium-dependent manner) [@liMolecularMechanismRegulate2016], and JIP3 and JIP4, which are involved in amino-acid induced lysosomal transport through TMEM55B [@willettTFEBRegulatesLysosomal2017][@gowrishankarOverlappingRolesJIP32021].  This is not an exhaustive list, and a number of other proteins have also been implicated in lysosome or late endosome transport in various different cell types.  The diversity of effector proteins involved in this process likely reflects the importance of lysosome regulation on the function of different tissues and cells, although further work in this area is required to build up a clearer picture.

#### Protrudin

As previously discussed (CROSS-REFERENCE TO INTRO SECTION), Raiborg et al. [@raiborgRepeatedEREndosome2015] proposed a mechanism through which protrudin promotes anterograde late endosome transport.  Subsequent work expanded this mechanism to lysosomes, and particularly implicated protrudin-mediated lysosomal transport in mTORC1 activation, albeit in non-neuronal cells [@hongPtdIns3PControlsMTORC12017].  

Given the importance of lysosome transport to cell growth and neuronal function, I decided to test whether protrudin overexpression could be increasing axonal lysosome transport.  


## Methods

General cell culture and microscopy methods can be found in section \@ref(METHODS).  Methods for the experiment presented in section \@ref(protrudin-ONC) are published in [@petrovaProtrudinFunctionsEndoplasmic2020].

### Cloning of LAMP1 construct for lysosome^[While LAMP1 labels a variety of lysosome-related organelles [@chengRevisitingLAMP1Marker2018], ranging from late endosomes to degradative autophagosomes, I will refer to LAMP1-labelled organelles as lysosomes throughout this thesis for the sake of brevity.] labelling

The LAMP1-RFP plasmid *(Addgene, 1817)* was modified to replace the existing CMV promoter with a human synapsin promoter to improve expression level in neurons.  Primers (see table \@ref(tab:primer-table)) were designed using the NEBuilder tool (https://nebuilder.neb.com) from New England Biolabs. 

```{r primer-table, echo = FALSE}
kable(#TODO - table styles
  data.frame(
    row.names = c('plasmid forward', 'plasmid reverse', 'insert forward', 'insert reverse'),
    sequence = c('GGTTTAGTGAACCGTCAG', 'CGGAACTCCATATATGGG', toupper('agcccatatatggagttccgCTAGACTCTAGCTGCAGAGGG'), toupper('atctgacggttcactaaaccTCGACTGCGCTCTCAGGC'))
  ),
  booktabs = TRUE,
  caption = 'Gibson cloning primers.'
  )
```

Fragments for cloning were produced by PCR, separated on a 1% agarose TBE gel, and purified using the Wizard SV Gel and PCR Clean-Up System kit *(Promega, A9282)*. 20$\mu$L assembly reactions were set up with Gibson master mix *(New England Biolabs, E2611S)* and an appropriate ratio of insert and vector DNA (2:1 or higher) and incubated at 50$^{\circ}$ for 1.5 hours. Assembled constructs were scaled up and successful cloning confirmed with restriction digest and DNA sequencing.

### Calcyon shRNA silencing
Ready-cloned shRNA expression vectors against calcyon were obtained from OriGene *(TG712969)* (see table \@ref(tab:caly-tab)).  The constructs readily expressed in rat cortical neurons (as seen from TurboGFP signal), but proved highly toxic to the cells.  Many cells transfected with either a mixture of Caly shRNA constructs or the scrambled control shRNA *(Origene, TR30013)* showed complete fragmentation, and the remaining intact cells had abnormally low levels of organelle movement, suggesting low viability.  This experiment was not continued further.

```{r caly-tab, echo = FALSE}
kable(#TODO - table styles
  data.frame(
    row.names = c('TG712969A', 'TG712969B', 'TG712969C', 'TG712969D'),
    sequence = c('ATCCGCAGAAGGCGGAGGATGTGCCGTCC', 
                 'TGCGGACCAGCCAAAGAAGTTCGCAGATT', 
                 'CCAGATGGCTTCCTACTTCGGCACAAGAT',
                 'TCAGCCATCATTCCCTGACCAGGTGGTCA')
  ),
  booktabs = TRUE,
  caption = 'Calcyon shRNA sequences.'
  )
```

### Lysotracker labelling
I tried labelling cells with LysoTracker Red *(Thermo Fisher, L7528)*, using the manufacturer's labelling protocol.  However, the dye was toxic to our neuronal cultures, even at low concentrations.  In addition, the density of our cultures made it difficult to distinguish lysosomes in the neuron of interest from those in overlapping neuronal projections.  As a result, I decided to proceed with transfected fluorescently-tagged markers instead of LysoTracker or other cell-permeable dyes.


## Results

### Protrudin overexpression is neuroprotective following axon injury *in vivo*{#protrudin-ONC}

*The data in this section has been published in [@petrovaProtrudinFunctionsEndoplasmic2020]*

Previous experiments in our lab had shown that protrudin overexpression improved the survival of cortical neurons subject to laser axotomy, in addition to promoting axon regeneration.  Therefore, we wanted to test if this neuroprotective effect could be replicated *in vivo*.  

Mice received intravitreal injections of AAV2 protrudin constructs, followed two weeks later by a unilateral optic nerve crush.  Two weeks post-crush, the mice were perfused, and eyes and optic nerves were collected.  RBPMS staining of retinal flat mounts showed a significant increase in RGC survival with expression of the active protrudin construct (Figure \@ref(fig:ONC_figure))[@petrovaProtrudinFunctionsEndoplasmic2020].  

```{r ONC_figure, echo = F, fig.path = '../figures/full/', fig.width = 16, fig.height = 6, fig.cap='Reproduced from Petrova et al. 2020.  A) Representative images of RBPMS-stained retinal whole mounts.  B) Quantification of surviving RGCs.'}
#TODO - detail on which eye was injected and virus structure.  Plasmid map?

ONC_figure
```

This effect was further substantiated by mouse retinal explant experiments, which showed that transfection with either WT or active protrudin completely preserved RGC survival over three days *ex vivo*, comapred to a GFP control which lost approximately 55% of RGCs over the same time. [@petrovaProtrudinFunctionsEndoplasmic2020].

### Protrudin overexpression does not increase lysosome transport in the distal axon

While the effect of protrudin on neuronal survival is exciting, it is also puzzling, as it is not fully explained by an increase in axonal Rab11/integrin.  Therefore, I set out to see if protrudin could be increasing the anterograde transport of late endosomes or lysosomes into the distal axon, which could have neuroprotective as well as regenerative effects.  Cortical neurons were transfected to co-express LAMP1-mRFP1 and either EGFP or EGFP-tagged protrudin, and LAMP1 transport in the distal axon was imaged using spinning disc miscroscopy.  These videos were then used to generate kymographs (as described in \@ref(METHODS)), which were manually traced to give measures of lysosome number, run length, and transport speed.  Overexpression of either wild type or active protrudin had no significant effect on the number, distance, or direction of travel of lysosomes in the distal axon.  A very slight decrease in anterograde speed was observed in cells expressing active protrudin.

```{r LAMP1_full_figure, echo = F, fig.width = 16, fig.height = 8, fig.path = '../figures/full/', fig.cap="A) Representative axon image and kymographs showing LAMP1 transport in the distal axon of cortical neurons expressing either EGFP (control) or EGFP-tagged protrudin constructs.  B) Total number of LAMP1-labelled organelles in each cell imaged.  C) Average run length (per cell) of moving lysosomes.  Anterograde and retrograde were defined as continuous movements of at least 5μm away from or towards the cell body, respectively.  D) Average lysosome velocity (per cell).  P-values are from Kruskal-Wallis test, followed by Dunn's post-hoc test with Bonfferoni correction for multiple comparisions."}

LAMP1_full
```

The experiment was then repeated using Rab7 instead of LAMP1 as a more specific marker of late endosomes.  As the mechanism set out in [@raiborgRepeatedEREndosome2015] suggests that Rab7 is directly involved in protrudin-mediated transport of late endosomes, this should label only the exact population of organelles that are involved in this process.  While overexpression of Rab7 could potentially amplify any effect of protrudin overexpression on this system, this effect should be consistent between control and protrudin conditions.  Cortical neurons were co-transfected with mCherry of mCherry-tagged protrudin and Rab7a-mEmerald, and Rab7 transport imaged and analysed as described previously.  However, this showed no significant effect of protrudin on any measured parameter of lysosome transport.


```{r Rab7_full_figure, echo = F, fig.width = 16, fig.height = 8, fig.path = '../figures/full/', fig.cap="A) Representative axon image and kymographs showing Rab7 transport in the distal axon of cortical neurons expressing either mCherry (control) or mCherry-tagged protrudin constructs.  B) Total number of Rab7-labelled organelles in each cell imaged.  C) Average run length (per cell) of moving late endosomes.  Anterograde and retrograde were defined as continuous movements of at least 5μm away from or towards the cell body, respectively.  D) Average late endosome velocity (per cell).  P-values are from Kruskal-Wallis test, followed by Dunn's post-hoc test with Bonfferoni correction for multiple comparisions."}

Rab7_full
```




## Discussion
Somewhat surprisingly, we found that protrudin overexpression had very little if any effect on lysosomal and late endosomal transport in cortical neurons.  However, further examination of the literature provides a likely explanation.  The mechanism described in [@raiborgRepeatedEREndosome2015] - which used PC12 cells as a model system - relies on the protein FYCO1 as an adaptor to link late endosomes to kinesin for anterograde transport.  This protein, also known as ZFYVE7, contains similar Rab7- and PIP-, and kinesin-binding domains to protrudin.  Raiborg et al. suggest that protrudin brings late endosomes and kinesin together through Rab7 and PI(3)P binding, but then transfers the kinesin onto Rab7-bound FYCO1 to facilitate transport.   However, transcriptomic data from [@kosekiSelectiveRab11Transport2017] suggests that FYCO1 is mostly absent from mature rat cortical neurons, which could explain why we see no effect on late endosome transport with protrudin overexpression.  This is consistent with transcriptomic data from other studies that show extremely low FYCO1 in human cortical neurons, expression biased towards non-neuronal cell types in the brain, and low levels in mouse DRG neurons [@zhangPurificationCharacterizationProgenitor2016][@kelleyVariationIntactTissue2018][@tedeschiCalciumChannelSubunit2016].

```{r FYCO1_RNAseq_full_figuree, echo = F, fig.width = 16, fig.height=12, fig.path='../figures/full/', fig.cap='TODO'}

FYCO1_RNAseq_full
```


Of note, [@ozkanERLysosomeContacts2021] also do not find a link between protrudin and CNS axonal lysosome transport.  They knocked down protrudin in hippocampal neurons, but found no effect on the distribution of LAMP1-labelled lysosomes into the axon.   This supports the idea that protrudin is not involved with late endosome/lysosome transport in neurons, but that other transport mechanisms (e.g. BORC) are predominant in axon transport.

While these findings can effectively rule out late endosome/lysosome transport as a mechanism through which protrudin drives neuroprotection and axon regeneration, they are worth bearing in mind when it comes to clinical translation.  The differential expression of FYCO1 across cell types suggests potential variations in the regulation of lysosome transport between them.  While protrudin overexpression has no substantial effect on this transport in cortical neurons, it could have an impact on other populations, leading to possible off-target effects in non-neuronal cell types.  Any gene therapies incorporating protrudin should therefore be carefully targeted to limit expression in the PNS and glial populations.

It remains an open question whether or not enhancing axonal lysosome transport could promote axon regeneration, for many of the reasons discussed at the start of this chapter.  Further research in this area is required, either targeting transport mechanisms such as BORC or Caly (which have been demonstrated to act in CNS neurons), or through exogenous expression of FYCO1.  In fact, [@saridakiFYCO1MediatesClearance2018] have already shown that FYCO1 overexpression in a *Drosophila* model of Parkinson's disease preserves motor function at baseline levels and prevents age-related decline.  (That being said, *Drosophila* does not contain an ortholog for protrudin, so FYCO1 may not have the same effect as it would in a vertebrate model [@wangHereditarySpasticParaplegia2008].)  While FYCO1 is an appealing candidate for further study, due to the possiblity of it acting synergistically with protrudin, it might be unfeasible as a gene therapy due to the size of the FYCO1 transcript.  At over 4.4kbp, it is very close to the packaging capacity of most AAV capsids, even without a promoter or other vector elements.  However, a better understanding of FYCO1 function might allow for the development of a shorter functional form with better translational potential.


