---
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
bibliography: "`r here::here('bib/THESIS.bib')`"
always_allow_html: true
---

# Is protrudin's effect on neuroprotection mediated by lysosome and/or late endosome transport?

```{=latex}
\chaptermark{Lysosome and late endosome transport}
```

```{r chapter_1_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

knitr::knit_child(here::here("analysis/transfection_optimisation.Rmd"))
knitr::knit_child(here::here("analysis/ONC_figure.Rmd"))
knitr::knit_child(here::here("analysis/Ptdn_LAMP1_analysis.Rmd"))
knitr::knit_child(here::here("analysis/Ptdn_Rab7_analysis.Rmd"))
knitr::knit_child(here::here("analysis/FYCO1_RNAseq_analysis.Rmd"))

source(here::here('style/thesis_setup.R'))

library(knitr)
library(here)
library(readbitmap)
library(kableExtra)
library(patchwork)

options(knitr.table.format = 'latex') #needed for scale down c.f. 'markdown'

```

## Introduction

### The many roles of neuronal lysosomes

Lysosomes have traditionally been regarded as the degradative compartment of the cell, responsible for the final breakdown and recycling of excess or damaged proteins and organelles [@deduveLysosome1963].  While this is certainly true, more recent work has shown that they are in fact a diverse pool of complex, multi-functional organelles.  While lysosomal function is essential to cellular survival broadly, it is particularly critical in neurons, which have additional metabolic, signalling, and transport constraints imposed by their extreme morphology.  As a result, normal activity and transport of lysosomes is a key factor in neuronal survival, and disruption of either of these processes can lead to neurodegeneration [@fergusonNeuronalLysosomes2019].

Lysosomal dysfunction is implicated in a wide range of neurodegenerative diseases, including Alzheimer's disease, Parkinson's disease, and ALS.  In many cases, defects in lysosomal degradation of autophagosome contents (a process which occurs in tandem with retrograde transport in axons [reviewed in @fergusonNeuronalLysosomes2019]) are thought to cause the accumulation of damaged organelles, which eventually leads to axonal swellings and subsequent degeneration [@dehayPathogenicLysosomalDepletion2010; @leeLysosomalProteolysisInhibition2011; @xieEndolysosomalDeficitsAugment2015].  Other conditions have been linked to impaired transport leading to decreased availability of lysosomes in the axon [@roneyLipidmediatedMotoradaptorSequestration2021].  Understanding how lysosomes function in neurons and the processes controlling their transport is therefore an important step in developing new protective or regenerative therapies for the CNS.

#### RNA transport

In recent years, a substantial body of evidence has emerged to demonstrate how lysosomes and closely related late endosomes facilitate RNA transport in neurons.  mRNA transport, and subsequent local protein synthesis, is essential for axon growth, pathfinding, maintenance, and synapse formation [reviewed in @cioniMolecularControlLocal2018].  

Cioni et al. demonstrated that ribonucleoproteins in the axon often associate with RAB7-marked late endosomes, and that these endosomes act as sites for local translation [@cioniLateEndosomesAct2019].  Furthermore, these late endosomes can form contact sites with mitochondria, where they are involved in the local translation of proteins for mitochondrial maintenance.  Work by Liao et al. identified the protein ANXA11 as the tether between lysosomes and RNA granules in the axon, and also showed that ALS-associated mutations in ANXA11 interfere with lysosome binding [@liaoRNAGranulesHitchhike2019].  

#### Signalling platform

As an endpoint of several protein degradation pathways, lysosomes are particularly responsive to amino acid levels in the cell, and in turn regulate the cell's response to feeding or starvation through signalling complex mTORC1.  mTORC1 is a serine/threonine kinase that localises to lysosomes, where it can be activated by regulatory proteins that colocalise in the presence of amino acids, in turn leading to increased protein translation and cell growth [reviewed in @inpanathanLysosomeSignalingPlatform2019].  This process is modulated by lysosome positioning — during starvation, lysosomes cluster near the nucleus, and mTORC1 activity is inhibited.  Feeding allows the lysosomes to transport towards the cell periphery, where mTORC1 activation can promote growth.  Hong et al. found that protrudin was involved in this process, interacting with lysosomes to facilitate anterograde transport [@hongPtdIns3PControlsMTORC12017].   This signalling is especially important in neurons — Poulopoulos et al. found that the majority of mRNA transcripts that they identified as enriched in growth cones contain a TOP motif for mTORC1-dependent translation.  In addition, they showed that endogenous mTOR and mTORC1 proteins were also enriched in growth cones, providing the necessary machinery for local translation to occur [@poulopoulosSubcellularTranscriptomesProteomes2019].  Therefore, the location of lysosomes and the regulation of mTORC1 on those lysosomes likely acts as a key gatekeeper of local translation, and the associated axonal functions that depend on it.

```{r mTORC1-schematic, echo = F, fig.path = '../figures/full/', fig.width = 16, fig.height = 20, fig.cap = 'A highly simplified schematic showing mTORC1 activation through multiple pathways in the presence of amino acids.  When amino acids are present (A), GATOR2 inhibits GATOR1, which in turn allows the RAG GTPase heterodimers to recruit the mTORC1 complex, which is further actiavted by amino acids and RHEB GTPase.  When amino acids are scarce (B), inhibition from GATOR1 prevents the mTORC1 recruitment by the RAGs.', fig.scap= 'mTORC1 activation schematic'}

image_as_ggplot(read.bitmap(here('figures/external/mTORC1_aa.png')))/
image_as_ggplot(read.bitmap(here('figures/external/mTORC1_no-aa.png'))) + plot_annotation(tag_levels = 'A')

```

Beyond mTORC1, lysosomes also participate in calcium signalling and homeostasis, which are particularly important for neuronal function.  Lysosomes are known to act as calcium stores for the cell, but also participate in a process known as store-operated calcium entry in response to calcium influx [@huiReleaseCalciumEndolysosomes2015].  In this process, lysosomes are exocytosed, inserting N-type calcium channels from their membranes into the plasma membrane and allowing increased calcium entry and subsequent activation of signalling cascades.  This process is required for a number of neuron-specific cellular functions, including electrical signalling, learning, and dendritic spine maintenance [reviewed in @wegierskiNeuronalCalciumSignaling2018].  Additionally, the exocytosis of lysosomes releases cathepsin B into the extracellular space, which activates matrix metalloproteinase 9 (MMP-9) through cleaving its inhibitor TIMP-1.  Active MMP-9 breaks down components of the extracellular matrix, facilitating cellular growth.  In neurons, this allows dendritic spines to grow in response to activity [@padamseyActivityDependentExocytosisLysosomes2017].  In cancer cells, this underpins the growth of invasive protrusions called invadopodia, a process that has been shown to depend on protrudin, RAB7, and FYCO1 [@pedersenProtrudinmediatedEREndosome2020]. 

#### ER morphology

Recent advances in microscopy have allowed the interactions between lysosomes and the ER to be imaged in unprecedented detail, and in turn revealed how lysosomes are involved in shaping the tubular ER network in cells.   Lu et al. demonstrated that, as lysosomes moved in response to the cell's nutritional state, they pulled ER tubules with them [@luStructureGlobalDistribution2020].  When a lysosome 'detached' from a contact site with the ER, the lysosome speed increased while the associated ER tubule retracted, indicating that the lysosomes do indeed pull on ER tubules rather than vice versa, or another mechanism of co-transport.  Interfering with the anterograde movement of lysosomes or ER-lysosomal contact sites resulted in fewer peripheral tubules and large ER sheets near the nucleus.  In neurons, this led to decreased axonal ER tubules and stunted axon growth.  Spits et al. showed a similar interaction with late endosomes, where RAB7 and its transport adaptors mediated ER tubule reorganisation, and resulted in a more sparse ER network in response to starvation [@spitsMobileLateEndosomes2021].   

As much as lysosomes help shape the ER, these interactions are reciprocal, and also play an essential role in lysosome morphology.  Disruption of the ER tubule network has been shown to result in the accumulation of large, static lysosomes near the nucleus, and prevent lysosome transport into axons [@ozkanERLysosomeContacts2021].  Other work has shown that association with ER tubules is necessary for lysosomal tubulation and cargo sorting, and that a key protein involved, LRRK2, is implicated in Parkinson's disease [@bonet-ponceEndoplasmicReticulumContributes2022].

### Lysosome transport mechanisms

As lysosomal function is closely tied to the positioning of these organelles in the cell, significant work has gone into understanding how lysosomes are transported.  This has identified a number of distinct, sometimes overlapping pathways that are regulated in different ways.  Work is still ongoing to understand the importance of these different pathways, and how they may interact in different cell types to control late endosome and lysosomal transport.

#### BORC/ARL8B/SKIP

The BORC/ARL8B/SKIP interaction is one of the best-characterised mechanisms involved in lysosome transport.  In summary, BORC, a complex of eight proteins, is tethered to the lysosomal membrane by a myristoyl tail on subunit protein myrlysin [@puRagulatorBORCInteraction2017].  BORC recruits and activates ARL8B, a small GTPase, which is otherwise diffusely distributed around the cell.  GTP-bound ARL8B can in turn bind to SKIP, which binds to kinesin-1.  This chain of interactions tethers lysosomes to the kinesin motor for transport along microtubules [@guardiaBORCFunctionsUpstream2016]. BORC activity is regulated by interaction with the Ragulator complex, which conveys amino acid sensitivity and allows for control of lysosomal positioning in response to nutrient levels in the cell [@puRagulatorBORCInteraction2017].  This process has been shown to mediate anterograde transport of lysosomes in axons [@fariasBORCKinesin1Ensemble2017], and defects in BORC component diaskedin are linked to progressive axonal dystrophy [@snouwaertMutationBorcs7Subunit2018].  In cancer cells, increased expression of several BORC subunits in cells that survived radiation therapy was shown to enhance cell invasiveness, supporting a role for BORC-mediated lysosome transport in cell outgrowth [@wuLysosomalTraffickingMediated2020].

\vspace{30pt}
```{r BORC-schematic, echo = F, fig.path = '../figures/full/', fig.width = 16, fig.height = 8, fig.cap = 'A schamatic of the interactions between BORC, ARL8B, SKIP, and kinesin, which allow lysosomes to be anterogradely transported along microtubules.', fig.scap= 'BORC schematic'}

image_as_ggplot(read.bitmap(here('figures/external/BORC_schematic.png')))

```

#### Calcyon and other adaptor proteins

While the BORC complex is involved in kinesin-mediated transport, several other adaptor proteins facilitate minus-end directed transport through binding to dyneins.  These include calcyon (which binds to late endosomes and lysosomes) [@shiDyneinBindsStimulates2017; @shiCouplingMicrotubuleMotors2018], RILP (which is a specific adaptor for RAB7)[@bucciExpressionAnalysisChromosomal2001], ALG2 (which binds lysosomes in a calcium-dependent manner) [@liMolecularMechanismRegulate2016], and JIP3 and JIP4, which are involved in amino-acid induced lysosomal transport through TMEM55B [@willettTFEBRegulatesLysosomal2017; @gowrishankarOverlappingRolesJIP32021].  The diversity of effector proteins involved in this process likely reflects the importance of lysosome positioning on the function of diverse tissues and cell types.

#### Protrudin

As previously discussed (section \@ref(intro-ptdn-transport)), Raiborg et al. proposed a mechanism through which protrudin promotes anterograde late endosome transport [@raiborgRepeatedEREndosome2015], which in turn drives neurite outgrowth.  Subsequent work expanded this mechanism to lysosomes, and particularly implicated protrudin-mediated lysosomal transport in mTORC1 activation, albeit in non-neuronal RPE-1 and HEK293 cells [@hongPtdIns3PControlsMTORC12017].  Additionally, since Petrova et al. showed that overexpression of protrudin increased the amount of ER in growth cones, it seemed likely that lysosomal transport might play a role in this process, in keeping with the role of lysosomes in ER transport discussed in section \@ref(intro-ptdn-axonalER) [@petrovaProtrudinFunctionsEndoplasmic2020].

### Hypothesis and experimental aims 

Given the importance of lysosome transport to cell growth and neuronal function, we decided to test whether the neuroprotective and pro-regenerative effects of protrudin overexpression could be mediated through an increase in axonal lysosome transport.  Previous work in the lab showed that RAB11 endosomes — with their cargo of integrins, growth factors, and other growth-related molecules — are excluded from CNS axons as they mature [@kosekiSelectiveRab11Transport2017].  Overexpression of protrudin can increase the amount of RAB11 reaching the distal axon, supporting regeneration [@petrovaProtrudinFunctionsEndoplasmic2020].  We reasoned that protrudin might act in a similar way on lysosomal transport, and that an increased number of lysosomes in the mature axon might promote axon survival and regrowth after injury.   To that end, we examined the effect of protrudin overexpression on axonal lysosome and late endosome numbers, speed, and other transport parameters.  At the same time, we set out to test whether other approaches to increase lysosomal transport might also promote regeneration or survival in damaged axons.



## Methods

General cell culture and microscopy methods can be found in section \@ref(METHODS).  Methods for the experiment presented in section \@ref(protrudin-ONC) are published in [@petrovaProtrudinFunctionsEndoplasmic2020].

### Cloning of LAMP1 construct for lysosome^[While LAMP1 labels a variety of lysosome-related organelles [@chengRevisitingLAMP1Marker2018], ranging from late endosomes to degradative autophagosomes, I will refer to LAMP1-labelled organelles as lysosomes throughout this thesis for the sake of brevity.] labelling{#cloning}

The LAMP1-RFP plasmid *(Addgene, 1817)* was modified to replace the existing CMV promoter with a human^[We use the human synapsin promoter in our lab instead of the rat synapsin promoter because it reliably drives expression in both human and rodent neurons, and is more suitable for clinical translation into human gene therapies.] synapsin promoter to improve expression level in neurons.  Primers (table \@ref(tab:primer-table)) were designed using the NEBuilder tool (`https://nebuilder.neb.com`) from New England Biolabs. 

Fragments for cloning were produced by PCR, separated on a 1% agarose TBE gel, and purified using the Wizard SV Gel and PCR Clean-Up System kit *(Promega, A9282)*. 20 \textmu{}L assembly reactions were set up with Gibson master mix *(New England Biolabs, E2611S)* and an appropriate ratio of insert and vector DNA (2:1 or higher) and incubated at 50°C for 1.5 hours. Assembled constructs were scaled up, and successful cloning confirmed with restriction digest and DNA sequencing.

```{r primer-table, echo = F, fig.pos='h'}
kable(
  data.frame(
    row.names = c('plasmid forward', 'plasmid reverse', 'insert forward', 'insert reverse'),
    sequence = c('GGTTTAGTGAACCGTCAG', 'CGGAACTCCATATATGGG', toupper('agcccatatatggagttccgCTAGACTCTAGCTGCAGAGGG'), toupper('atctgacggttcactaaaccTCGACTGCGCTCTCAGGC'))
  ),
  booktabs = TRUE,
  caption = 'Gibson cloning primers.'
  ) %>% 
  kable_styling(latex_options="scale_down")

```

### Calcyon shRNA silencing

Ready-cloned shRNA expression vectors against calcyon were obtained from OriGene *(TG712969)* (table \@ref(tab:caly-tab)).  The constructs readily expressed in rat cortical neurons (as seen from TurboGFP signal), but proved highly toxic to the cells.  Many cells transfected with either a mixture of calcyon shRNA constructs or the scrambled control shRNA *(Origene, TR30013)* showed complete fragmentation, and the remaining intact cells had abnormally low levels of organelle movement, suggesting low viability.  This experiment was not continued further.

```{r caly-tab, echo = FALSE, fig.pos='h'}
kable(
  data.frame(
    row.names = c('TG712969A', 'TG712969B', 'TG712969C', 'TG712969D'),
    sequence = c('ATCCGCAGAAGGCGGAGGATGTGCCGTCC', 
                 'TGCGGACCAGCCAAAGAAGTTCGCAGATT', 
                 'CCAGATGGCTTCCTACTTCGGCACAAGAT',
                 'TCAGCCATCATTCCCTGACCAGGTGGTCA')
  ),
  booktabs = TRUE,
  caption = 'Calcyon shRNA sequences.'
  )
```

### Lysotracker labelling

We initially tried labelling cells with LysoTracker Red *(Thermo Fisher, L7528)*, using the manufacturer's labelling protocol.  However, the dye was toxic to our neuronal cultures, even at low concentrations.  In addition, the density of our cultures made it difficult to distinguish lysosomes in the neuron of interest from those in overlapping neuronal projections.  As a result, we decided to proceed with transfected fluorescently-tagged markers instead of LysoTracker or other cell-permeable dyes.

### BORC-related constructs for lysosome/late endosome manipulation

Expression constructs for SKIP, myrlysin (BORCS5), WT ARL8B, and constitutively active ARL8B were kindly provided by Juan Bonifacino.  These constructs were scaled up as described in section \@ref(plasmid-prep) and tested for transfection in cortical neurons.  Additional constructs of SYN-myrlysin-EGFP and CBh-SKIP-mNeonGreen were obtained from VectorBuilder and prepared in the same way.

\vspace{14pt}
```{r SKIP-LAMP1-kymo, echo = F, fig.height=4, dpi = 160, fig.cap="Representative kymograph showing SKIP in green and LAMP1 in magenta.  Scale bar is 20 \\textmu{}m.", fig.scap="SKIP and LAMP1 representative kymograph"}
image_as_ggplot(read.bitmap(here('figures/external/SKIP_LAMP1_kymo_annotated.png')))
```

Our original intention was to test if artificially increasing lysosomal transport in the axon could promote regeneration; however, this work was interrupted by the pandemic.  Following our return to the laboratory, we conducted some pilot experiments with SKIP and LAMP1.  This showed some colocalisation between the two, particularly in anterogradely-moving lysosomes, but no immediately obvious differences in number of lysosomes, speed, or run length compared to an EGFP control (figure \@ref(fig:SKIP-LAMP1-kymo)).  (This is not unexpected, as previous experiments required both SKIP and ARL8B overexpression to stimulate transport [@rosa-ferreiraArl8SKIPAct2011; @fariasBORCKinesin1Ensemble2017].)  After the completion of the experiments described in section \@ref(lyso-LE), it was decided not to continue this work further.


### Transfection optimisation

As mentioned in section \@ref(METHODS), we experienced a sharp decline in efficiency with our initial magnetofection protocol, and needed to reoptimise our process.  This occurred between the experiments discussed in section \@ref(lyso-LE).  During this time, we tested a variety of protocols and modifications to identify the factors most important for transfection efficiency, using the same EGFP/Emerald expression construct.  Results were measured by scanning the entire plate of cells on a Leica DMi8 tilescan microscope and counting the number of transfected neurons that were either intact or visibly dead (figure \@ref(fig:transfection-plot-full)).  

```{r transfection-plot-full, echo = F, fig.pos = '!h', fig.width = 16, fig.height = 16, fig.cap="A) Example image of a full-plate scan used for quantifying transfection.  The indicated area (1x1 mm) is enlarged to the right, and examples of intact cells (blue asterisk) and dead cells (magenta x) are annotated. B) Cell counts for plates of varying density transfected with magnetofection protocol.  C) Cell counts for three identical plates (300,000 cells per plate) transfected with calcium phosphate protocol.  Note the difference in scale between the two plots.", fig.scap="Transfection optimisation plot"}

optimisation_plot
```

Of all the factors tested, cell density best correlated with the transfection efficiency, although the improvements on the original magnetofection protocol were dwarfed by the vast improvement with calcium phosphate transfection.  Going forward, transfections were carried out using the calcium phosphate method, with cell density of 200,000-400,000 per imaging dish depending on the needs of the experiment.


## Results^[The results presented in this section largely reflect my first year of study, a time when I was also training to perform *in vivo* experiments to assess neuroprotection and regeneration in the eye and optic nerve.  My initial plan was to determine protrudin's effect on lysosome/late endosome transport, and then test a variety of different tools (e.g. calcyon shRNA, BORC overexpression, etc.) to manipulate lysosome and late endosome transport and assess their regenerative potential.  Unfortunately, the cessation of lab work during the pandemic and associated disruptions eventually led us to abandon most of these plans.]

### Protrudin overexpression is neuroprotective following axon injury *in vivo*{#protrudin-ONC}

*The data in this section has been published in [@petrovaProtrudinFunctionsEndoplasmic2020].  My contribution to the experiment included DNA preparation, tissue collection, and data curation and analysis.*

The optic nerve crush model is a powerful tool for examining CNS survival and regeneration *in vivo*.  Because the eye and the optic nerve can be accessed much more readily than, for example, the brain and spinal cord, it allows the gene delivery and axon injury to be performed with minimal invasiveness.  In addition, because the bodies of the RGCs are all located together in the retina, it is possible to quantify both cell survival and axon regrowth in the same experiment.

For this experiment, mice received intravitreal injections of AAV2 viruses containing either EGFP only, WT protrudin, or active protrudin under a CAG promoter, followed two weeks later by a unilateral optic nerve crush.  Two weeks post-crush, the mice were perfused, and eyes and optic nerves were collected.  RBPMS staining of retinal flat mounts showed a significant increase in RGC survival with expression of the active protrudin construct, as measured by comparing the number of RBPMS positive cells between the crush and uninjured eyes (figure \@ref(fig:ONC-figure))[@petrovaProtrudinFunctionsEndoplasmic2020].  

(ref:ONC-cap) Reproduced with permission from Petrova et al. 2020.  A) Timeline of the optic nerve crush experiments.  (CTB injection was included for the purpose of tracing regenerating axons, but is not shown in this figure.)  B) Representative images of RBPMS-stained retinal whole mounts.  Scale bar is 100 \textmu{}m.  C) Quantification of surviving RGCs.   Fisher’s exact test with analysis of stack of *p* values and Bonferroni–Dunn multiple comparison test. Error bars represent mean ± SEM.
```{r ONC-figure, echo = F, fig.path = '../figures/full/', fig.width = 16, fig.height = 12, fig.cap='(ref:ONC-cap)', fig.scap="RGC neuroprotection experiment from Petrova et al. 2020"}

ONC_figure
```

This effect was further substantiated by mouse retinal explant experiments, which showed that transfection with either WT or active protrudin completely preserved RGC survival over three days *ex vivo*, compared to an EGFP control which lost approximately 55% of RGCs over the same time [@petrovaProtrudinFunctionsEndoplasmic2020].

### Protrudin overexpression does not increase lysosome or late endosome transport in the distal axon{#lyso-LE}

To test the hypothesis that protrudin might function in part by increasing the anterograde transport of late endosomes or lysosomes into the distal axon, we set out to measure the effects of protrudin overexpression on the axonal transport of these organelles.  As described above, there are a number of ways in which this increased transport could have neuroprotective or regenerative effects, and such a function would be in keeping with protrudin's described behaviour in the literature.  

Rat cortical neurons were transfected to co-express LAMP1-mRFP1 as a marker of lysosomes, with either EGFP or EGFP-tagged protrudin.  LAMP1 transport in the distal axon was imaged in three-minute segments using spinning disc microscopy.  These videos were then used to generate kymographs as described in \@ref(kymo) (figure \@ref(fig:LAMP1-full-figure), panel A), which were manually traced to give measures of lysosome number, run length, and transport speed.  Overexpression of either wild type or active protrudin had no significant effect on the number, distance, or direction of travel of lysosomes in the distal axon (figure \@ref(fig:LAMP1-full-figure), B-C).  A very slight decrease in anterograde speed was observed in cells expressing active protrudin (figure \@ref(fig:LAMP1-full-figure), panel D).

(ref:LAMP1-caption) A) Representative axon image and kymographs showing LAMP1 transport in the distal axon of cortical neurons expressing either EGFP (control) or EGFP-tagged protrudin constructs.  All scale bars are 20 \textmu{}m.  B) Total number of LAMP1-labelled organelles in each cell imaged.  C) Average run length (per cell) of moving lysosomes.  D) Average lysosome velocity (per cell).  *n* = 28 to 30 cells per condition across four independent batches.  Kruskal-Wallis *p* values are shown at the top of each plot; significant *p* values from Dunn's post-hoc test with Bonferroni correction for multiple comparisons are shown over brackets.
```{r LAMP1-full-figure, echo = F, fig.width = 16, fig.height = 8, fig.path = '../figures/full/', fig.cap="(ref:LAMP1-caption)", fig.scap = "Protrudin effect on lysosome (LAMP1) transport in the distal axon"}

LAMP1_full
```

The experiment was then repeated using RAB7 instead of LAMP1 as a more specific marker of late endosomes.  As the mechanism set out in Raiborg et al. suggests that RAB7 is directly involved in protrudin-mediated transport of late endosomes, this should label only the population of organelles that are involved in this process [@raiborgRepeatedEREndosome2015].  While overexpression of RAB7 could potentially amplify any effect of protrudin overexpression on this system, this effect should be consistent between control and protrudin conditions.  Cortical neurons were co-transfected with mCherry or mCherry-tagged protrudin and RAB7A-mEmerald, and RAB7 transport imaged and analysed as described previously.  However, this showed no significant effect of protrudin on any measured parameter of transport (figure \@ref(fig:RAB7-full-figure)).  Importantly, there was no change in the total number of lysosomes or late endosomes with either WT or active protrudin expression, compared to a control (figure \@ref(fig:RAB7-full-figure), panel B).

(ref:RAB7-caption) A) Representative axon image and kymographs showing RAB7 transport in the distal axon of cortical neurons expressing either mCherry (control) or mCherry-tagged protrudin constructs.  All scale bars are 20 \textmu{}m.  B) Total number of RAB7-labelled organelles in each cell imaged.  C) Average run length (per cell) of moving late endosomes.  D) Average late endosome velocity (per cell).  *n* = 17 or 18 cells per condition across three independent batches.  Kruskal-Wallis *p* values are shown at the top of each plot.
```{r RAB7-full-figure, echo = F, fig.width = 16, fig.height = 8, fig.path = '../figures/full/', fig.cap="(ref:RAB7-caption)", fig.scap = "Protrudin effect on late endosome (RAB7) transport in the distal axon"}

RAB7_full
```

Taken together, these results demonstrate that neither wild type nor active protrudin has any clear effect on lysosome or late endosome transport in the axon, and do not support our initial hypothesis that protrudin's beneficial effects are associated with an increase in transport of these organelles.

## Discussion{#no-FYCO}

Somewhat surprisingly, we found that protrudin overexpression had little if any effect on lysosomal and late endosomal transport in cortical neurons.  However, further examination of the literature provides a likely explanation.  The mechanism described in Raiborg et al. — which used PC12 cells as a model system — relies on the protein FYCO1 as an adaptor to link late endosomes to kinesin for anterograde transport [@raiborgRepeatedEREndosome2015].  This protein, also known as ZFYVE7, contains similar RAB7-, PIP-, and kinesin-binding domains to protrudin.  Raiborg et al. reported that protrudin brings late endosomes and kinesin together through RAB7 and PI(3)P binding, but then transfers the kinesin onto RAB7-bound FYCO1 to facilitate transport.   However, transcriptomic data from Koseki et al. suggests that FYCO1 mRNA is almost undetectable in cortical neurons, which could explain why we see no effect on late endosome transport with protrudin overexpression (figure \@ref(fig:FYCO1-RNAseq-full-figure), panel A) [@kosekiSelectiveRab11Transport2017].  This is consistent with transcriptomic data from other studies that show extremely low FYCO1 in human cortical neurons, expression biased towards non-neuronal cell types in the brain, and low levels in mouse DRG neurons, as illustrated in figure \@ref(fig:FYCO1-RNAseq-full-figure), B-G [@zhangPurificationCharacterizationProgenitor2016; @kelleyVariationIntactTissue2018; @tedeschiCalciumChannelSubunit2016].

The slight decrease in anterograde lysosome speed seen in figure \@ref(fig:LAMP1-full-figure), panel D, with the expression of active protrudin is unlikely to represent a direct effect on late endosome/lysosome transport, given the absence of any similar effect on RAB7 transport (figure \@ref(fig:RAB7-full-figure)).  This might instead indicate an interaction between LAMP1-labelled lysosomes and other organelles in the axon, directly or indirectly mediated by protrudin.  It could also potentially represent an increase in lysosomal size, perhaps due to an upstream effect on autophagy, which could result in slower axonal transport [@wangRadialContractilityActomyosin2020].  (The video and kymograph data of LAMP1 transport did not show an obvious change in lysosomal size, but this was not examined quantitatively, and any changes could be quite small given the magnitude of the effect on transport speed.)  I discuss these possibilities further in subsequent chapters.

(ref:FYCO-cap) RNA sequencing data from A) Koseki et al. 2017 [@kosekiSelectiveRab11Transport2017], B) Zhang et al. 2016 [@zhangPurificationCharacterizationProgenitor2016], C) Kelley et al. 2018 [@kelleyVariationIntactTissue2018], D) Tian et al. 2019 (via `https://ineuronrnaseq.shinyapps.io/rnaseq_app`) [@tianCRISPRInterferenceBasedPlatform2019], E-G) Tedeschi et al. 2016 [@tedeschiCalciumChannelSubunit2016].
\vspace{30pt}
```{r FYCO1-RNAseq-full-figure, echo = F, fig.width = 16, fig.height=16, fig.path='../figures/full/', fig.cap='(ref:FYCO-cap)', fig.scap="FYCO1 transcriptome comparisons"}

FYCO1_RNAseq_full
```

Of note, Özkan et al. also do not find a link between protrudin and CNS axonal lysosome transport.  They knocked down protrudin in hippocampal neurons, but found no effect on the distribution of LAMP1-labelled lysosomes into the axon  [@ozkanERLysosomeContacts2021].   This supports the idea that protrudin is not involved with late endosome/lysosome transport in CNS neurons, but that other transport mechanisms (e.g. BORC) are predominant in axon transport.

It is worth noting that these results do not contradict our earlier findings published Petrova et al. [@petrovaProtrudinFunctionsEndoplasmic2020].  The mechanism of late endosome transport set out in Raiborg et al. — in which protrudin interacts only transiently with RAB7 before passing the late endosome to FYCO1 — is substantially different from how it functions in recycling endosome transport.  In the case of recycling endosomes, protrudin is believed to act as the primary adaptor between RAB11 and kinesin, and therefore does not require FYCO1.

While these findings can effectively rule out late endosome/lysosome transport as a mechanism through which protrudin drives neuroprotection and axon regeneration, they are worth bearing in mind when it comes to clinical translation.  The differential expression of FYCO1 across cell types suggests potential variations in the regulation of lysosome transport between them.  While protrudin overexpression has no substantial effect on this transport in cortical neurons, it could have an impact on other populations, leading to possible undesired effects in non-neuronal cell types.  Any gene therapies incorporating protrudin should therefore be carefully targeted to limit expression in the PNS and glial populations.

It remains an open question whether or not enhancing axonal lysosome transport could promote axon regeneration, for many of the reasons discussed at the start of this chapter.  Further research in this area is required, either targeting transport mechanisms such as BORC or calcyon (which have been demonstrated to act in CNS neurons), or through exogenous expression of FYCO1.  In fact, Saridaki et al. have already shown that FYCO1 overexpression in a *Drosophila* model of Parkinson's disease preserves motor function at baseline levels and prevents age-related decline  [@saridakiFYCO1MediatesClearance2018].  (That being said, *Drosophila* does not contain an ortholog for protrudin, so FYCO1 may not have the same effect as it would in a vertebrate model [@wangHereditarySpasticParaplegia2008].)  While FYCO1 is an appealing candidate for further study, due to the possibility of it acting synergistically with protrudin, it might be unfeasible as a gene therapy due to the size of the FYCO1 transcript.  At over 4400 base pairs, it is very close to the packaging capacity of most AAV capsids, even without a promoter or other vector elements.  However, a better understanding of FYCO1 function might allow for the development of a shorter functional form with better translational potential.
