---
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
  word_document: default
bibliography: "`r here::here('bib/THESIS.bib')`"
always_allow_html: true
---

# ER, ERGIC, and Golgi in axons
```{=latex}
\chaptermark{ER, ERGIC, and Golgi}
```

```{r chapter_3_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r include=FALSE}
knitr::knit_child(here::here("analysis/ER-Golgi_localisation.Rmd"))
knitr::knit_child(here::here("analysis/Ptdn_St3Gal5_analysis.Rmd"))
knitr::knit_child(here::here("analysis/St3Gal5_puncta_counts.Rmd"))
knitr::knit_child(here::here("analysis/ESCargo_fig.Rmd"))
knitr::knit_child(here::here("analysis/St3Gal5_scratch.Rmd"))


library(knitr)

```

## Introduction

The discovery of local secretory pathways in neurons is fairly recent, but has already contributed significantly to our understanding of how these cells develop and function.  In non-polarised cells, protein translation and modification typically take place near the nucleus, and proteins are then transported to their final destination in, on, or outside of the cell.  While this does also occur in neurons, neurons are also capable of locally translating proteins, which can then be delivered to organelles or inserted into the cell membrane as needed [@meriandaAfunctionalequivalent2009].  It has become increasingly apparent that protein delivery from the cell body is not sufficient to maintain protein turnover at distal parts of the cell, nor is it fast enough to support the dynamic behaviour observed in these regions, for example during axon pathfinding.  Instead, these regions rely on several overlapping local pathways for protein production, modification, and secretion (reviewed in [@gonzalezGolgibypassreview2018]).  Recent work has even linked levels of local protein translation to axon regenerative capacity [@vanerpAgerelatedLossAxonal2021], indicating that this process is critically important to promoting CNS regeneration. However, while there is evidence for the activity of a local secretory pathway in dendrites and PNS axons, the same has not yet been demonstrated in CNS axons [@meriandaAfunctionalequivalent2009; @mikhaylovaGS2016; @cornejoNonconventionalAxonalOrganelles2020].

```{r local-schematic, echo = F, fig.width=10, fig.cap="Schematic showing key components of the local secretory pathway.", fig.scap="Schematic of the local secretory pathway", dpi = 300}
image_as_ggplot(read.bitmap(here('figures/external/local_secretory_pathway.png')))
```

### ER exit sites and COP vesicles

Key to all of these pathways is the presence of interconnected ER tubules throughout the neuron.  Continuous with the ER network in the soma, these tubules reach into dendrites, into individual spines in an activity-dependent manner [@perez-alvarezEndoplasmicReticulumVisits2020], and down the entirety of the axon.  Among other roles, this ER network provides a platform for local protein translation, and targeting to Golgi-related organelles for further modification.

Membrane proteins translated in the ER are exported via ER exit sites (ERES), typically in structures known as COPII-coated vesicles.  Recent work has demonstrated that these exit sites are complex tubular structures, formed by the interactions of numerous COPII-associated proteins [@weigelERtoGolgiProteinDelivery2021].  A number of different chaperones and cargo receptors are involved in this process, sorting proteins for export and facilitating the budding of these transport intermediates from the ER network.  Aridor and Fish showed that Sar1, a GTPase related to COPII budding, was targeted to mammalian axons during neuronal development, and that this was essential for axonal growth [@aridorSelectiveTargetingER2009].

### The ER-Golgi intermediate compartment

After budding from the ER, COPII-coated vesicles fuse into a dynamic structure known as the ER-Golgi intermediate compartment (ERGIC) (reviewed in [@appenzeller-herzogERGolgiIntermediateCompartment2006]).  From here, cargo is directed to different downstream organelles for further processing.

### Golgi outposts and satellites

The most common destination after leaving the ERGIC is the Golgi, although this can take different forms depending on the region of the cell involved.  In the cell body, this is typically the perinuclear Golgi stack.  In dendrites, proteins may be targeted to Golgi outposts or Golgi satellites.  While these are structurally distinct, they both carry out essential Golgi functions, including protein glycosylation [@mikhaylovaGS2016].  Recently, Golgi satellites have also been identified in axons, both in static and dynamically transported populations.  Here, they can act as local glycosylation platforms, similar to in dendrites, but also take part in delivering functional proteins to lysosomes [@liePostGolgicarriers2021] and nerve terminals [@cornejoNonconventionalAxonalOrganelles2020].  Golgi satellite localisation, and by extension their functionality, is dynamically regulated by neuronal activity [@govindActivitydependentGolgiSatellite2021].

### Further processing and protein secretion

After leaving a Golgi-related organelle, membrane proteins can be sorted in endosomes through the retromer complex, which returns cargo receptors to the Golgi.  This complex has been identified in dendrites closely associated with Golgi satellites, suggesting that this sorting also occurs as part of local secretory pathways [@mikhaylovaGS2016].  Recent work has linked retromer dysfunction to a number of neurodegenerative conditions, potentially through changes to the lysosomal proteome [@dalyMultiomicApproachCharacterises2022].

After sorting by the retromer complex, proteins can then be delivered to other organelles or the plasma membrane.  Several studies have found local translation, processing, and secretion of cell surface proteins, including voltage-gated sodium channels at nodes of Ranvier [@gonzalezAxonsprovidesecretory2016].  The exocyst complex facilitates fusion of post-Golgi carriers with the plasma membrane, and interference with this complex has been shown to inhibit neurite growth [@pereiraExocystComplexEssential2022; @swopeExocystComplexRequired2022].

### The Golgi-bypass pathway

An alternate, Golgi-bypassing pathway has also been suggested, with cargo moving from the ERGIC to the cell surface via Rab11-marked recycling endosomes.  This has been studied in dendrites, where it was found to play a role in delivering GluA1 glutamate receptors to dendritic spines [@bowenGolgiindependentSecretoryTrafficking2017].  This study does not fully rule out the involvement of Golgi satellites, but does show that this pathway is robust to the disruption of somatic Golgi and Golgi outposts with brefeldin A, indicating a distinction from conventional Golgi processing.  

## Methods

General cell culture and microscopy methods can be found in section \@ref(METHODS).

### i3 neuron culture

iPSC-derived i3 neurons were cultured as described in [@fernandopulleTranscriptionFactorMediated2018].  After three days of differentiation with doxycycline, neurons were plated on poly-L-ornithine *(Sigma-Aldrich, P3655-50MG)* coated coverslips and grown at 37°C with 7% CO~2~ in BrainPhys media *(STEMCELL Technologies, 05790)* supplemented with B27 *(Thermo Fisher, 17504044)*, BDNF *(PeproTech, 450-02)*, NT-3 *(PeproTech, 450-03)*, and laminin *(Thermo Fisher, 23017015)*.  

### Immunochemistry

Immunochemistry was carried out as described in section \@ref(immuno).  In the case of BCAP31, a biotinylated secondary antibody was added to improve signal, and cells were additionally stained for 1.5 hours at room temperature with dye-conjugated streptavidin *(Thermo Fisher, S11226)*.  All other steps were unchanged.

### Fluorescence quantification

Confocal images were analysed as maximum-intensity stack projections in Fiji.  Tracing of regions of interest was assisted by a custom macro, which is available by request.  The area of the nucleus was excluded from cell body quantification.  Background correction was carried out for axon terminals, axons, and dendrites by subtracting the intensity of an identical ROI, manually shifted to avoid overlap with the area of interest.  Cell bodies were not background-corrected, due to brighter overall intensity and the density of neurites and other cellular structures nearby.

### Inhibition of ER to Golgi transport

We tested several inhibitors of either ER to Golgi transport or Golgi function in DIV3 cortical neurons — H-89 (100$\mu$M) *(Calbiochem, 371962)*, clofibrate (500$\mu$M) *(Sigma-Aldrich, C6643)*, brefeldin A (10$\mu$g/mL) *(Calbiochem, 500583)*, and golgicide A *(Calbiochem, 345862)* (10$\mu$M).  These concentrations were based on previously published use [@helmsInhibitionBrefeldinGolgi1992; @defigueiredoClofibrateInhibitsMembrane1999a; @gonzalezAxonsprovidesecretory2016; @saenzGolgicideRevealsEssential2009; @weigelERtoGolgiProteinDelivery2021].  Concentration-matched DMSO-only controls were used for each inhibitor.  The inhibitors were added to cells for 30 minutes at 37°C, 7% CO~2~, in supplement-free media, after which cells were fixed.  Given protrudin's effect on ERGIC localisation reported below (figure \@ref(fig:ERGIC53-plot)), we used ERGIC53 staining as a measure of inhibitor function.  We found that H-89 caused growth cone collapse, while the others preserved cellular morphology, and clofibrate led to a slight decrease in ERGIC53 signal in both the cell body and growth cone (data not shown).

Based on these preliminary results, we decided to test if clofibrate could block protrudin's pro-regenerative effect on injured axons.  Unfortunately, this dosage of clofibrate proved highly toxic over the 16-hour timescale required for this experiment.  We were unable to optimise this experiment further, as our regular supply of rats for cortical neuron cultures was disrupted by an animal carriage embargo at Frankfurt airport during the summer of 2022.  We plan to return to this in a future project.

## Results

### ER-Golgi pathway localisation during development

We first set about to determine the baseline localisation of our selected markers, and to assess whether this changed during neuronal development, as regenerative ability declines.  We chose DIV3 (the earliest point at which axons can be clearly identified) and DIV15 (synaptic maturation and low regenerative ability) as developmental timepoints to analyse, based on previous work in the lab published in Koseki et al. [@@kosekiSelectiveRab11Transport2017]. We stained cortical neurons for BCAP31, SURF4, ERGIC53, and SLC38A1 at each timepoint, and compared fluorescence intensity between the cell body and axon terminal.  This showed that, with the exception of SURF4, all markers show a relative decline at the axon terminal with age (figure \@ref(fig:dev-changed-full-figure)).  Separate analysis of each region shows that all markers show increased intensity in the cell body with age, but that most show no significant change in the axon terminal, with the exception of SLC38A1, which increases (supplementary figure \@ref(fig:suppl-dev-grid)).  This is in keeping with SLC38A1's role as an amino acid transporter, which is a particularly important function for synaptically active neurons [@hellstenNeuronalAstrocyticProtein2017].  It is worth noting that, while the average intensity remains the same, the axon terminal dramatically reduces in size as the cell matures, so this represents a decrease in the total amount of these proteins present at the axon tip.

```{r dev-changed-full-figure, echo = F, fig.width=16, fig.height=16, fig.path='../figures/full/', fig.cap = "A) Ratio of fluorescence intensity between Cell body and axon terminal in DIV3 vs DIV15 neurons.  All n = 26 or 27 cells per condition across three batches, except SLC38A1 (n = 18 cells, three batches). B-M) Representative images of DIV3 neurons (B-E), DIV15 cell bodies (F-I), and DIV15 axon terminals (J-M)  All scale bars are 20 \\textmu{}m, and J-M are 20x20 \\textmu{}m.  Markers correspond to the graph at the top of each column.", fig.scap="Developmental changes in marker localisation in cortical neurons"}

dev_ratio_plot
```

The staining was repeated in DIV1 (post-differentiation) i3 neurons, at a stage when they are just beginning to grow axons, to confirm that the axon terminal:cell body ratio was conserved between rodent and human neurons.  This showed remarkable consistency between cell types, with the exception of BCAP31, which had a very slightly higher ratio in rat neurons.

```{r i3-ratio-plot, echo = F, fig.width=16, fig.height=10, fig.path='../figures/full/', fig.cap="A) Ratio of fluorescence intensity between Cell body and axon terminal in DIV1 i3 vs DIV3 rat cortical neurons.  n = 17 or 18 i3 neurons per condition across three batches. B-E) Representative images of DIV1 i3 neurons stained for each marker.  All scale bars are 20 \\textmu{}m.", fig.scap = "Comparison of marker localisation between i3 neurons and rat cortical neurons"}

i3_plot
```

We also tested a tool called ESCargo(FTV) to visualise ER exit sites [@caslerESC2020; @govindActivitydependentGolgiSatellite2021].  This encodes a SURF4-dependent synthetic secretory cargo, which can be inducibly released from the ER with the addition of a synthetic ligand.  While the aggregates were clearly visible in cell bodies, we also observed them in axons, although they were distributed very sparsely along their length (figure \@ref(fig:ESCargo-figure)).  In some cases, these were found at axonal swellings or branch points, raising the possibility that these exit sites may concentrate at particular structural features.  Further work is required to test if this corresponds to the localisation of functional ER exit sites in the axon, and whether their size or number is regulated developmentally or with the overexpression of protrudin.

```{r ESCargo-figure, echo = F, fig.width=16, fig.height=12, fig.path='../figures/full/', fig.cap="A) Cell body of a cell co-expressing EGFP/Emerald with ESCargo(FTV)-Crimson.  Note individual aggregates or small clusters in dendrites, often at branch points.  B) Axon of neuron depicted in A, with two ESCargo(FTV) aggregates circled.  All scale bars are 20 \\textmu{}m.", fig.scap="ESCargo localisation in cortical neurons"}

ESCargo_plot
```

### The role of protrudin in localisation and transport

Given the developmental decline in these markers at axon terminals, we were interested to see if protrudin might act by increasing the axonal levels of ER-Golgi pathway components.  We repeated the experiment above in DIV15 neurons expressing either EGFP or EGFP-tagged protrudin, and measured the fluorescence intensity in the cell body and axon terminals.

#### BCAP31

BCAP31 showed high variability between cells, particularly in the soma.  This may have been a result of signal amplification with a biotinylated secondary antibody, although the smaller range of intensities measured at axon terminals suggests that this may represent natural variation.  In any case, while protrudin did appear to decrease cell body BCAP31 levels, this did not translate to a change in the axon terminal:cell body ratio.  Therefore, we do not believe that protrudin substantially changes the distribution of BCAP31 in cells.  (Figure \@ref(fig:BCAP31-plot))

Peculiarly, in many cases BCAP31 signal appeared higher in cells expressing either protrudin or control EGFP, compared to untransfected neighbouring cells.  It is possible that the levels of this protein are dynamically controlled in response to the cell's overall level of protein synthesis, which may be impacted by exogenous plasmids.  This should be taken into account in any future experiments examining BCAP31 levels or localisation.

#### SURF4

SURF4 decreased both in the cell body and axon terminal of cells expressing active protrudin.  However, these decreases were well-matched, resulting in no change to the axon terminal:cell body ratio.  It is possible that protrudin is not changing the localisation of this ERES/ERGIC protein, but may alter its overall expression levels.  (Figure \@ref(fig:SURF4-plot))

#### SLC38A1

SLC38A1 showed no significant changes with either WT or active protrudin expression, although there was a slight trend towards higher axonal intensity.    (Figure \@ref(fig:SLC38A1-plot))

#### ERGIC53

In contrast with the other markers, ERGIC53 showed significantly altered localisation with the expression of active, but not wild type, protrudin.  This included a significant increase in axon terminal intensity and axon terminal:cell body ratio, on top of a slight increase in soma intensity.  This suggests that protrudin is playing a role in either the formation of ERGIC compartments, or their transport into the distal axon.  (Figure \@ref(fig:ERGIC53-plot))


```{r BCAP31-plot, echo = F, fig.width=16, fig.height=16, fig.path='../figures/full', fig.cap="A-F) Representative images of BCAP31 in control, WT, and active protrudin-expressing cell bodies (A-C, respectively) and axon terminals (D-F).  All scale bars are 20 \\textmu{}m, and D-F are 20x20 \\textmu{}m.  G-I) Quantification of BCAP31 fluorescence intensity. n = 18 cells per condition across three independent batches.", fig.scap="BCAP31 localisation in protrudin-expressing neurons"}

BCAP31_plot
```

```{r SURF4-plot, echo = F, fig.width=16, fig.height=16, fig.path='../figures/full', fig.cap="A-F) Representative images of SURF4 in control, WT, and active protrudin-expressing cell bodies (A-C, respectively) and axon terminals (D-F).  All scale bars are 20 \\textmu{}m, and D-F are 20x20 \\textmu{}m.  G-I) Quantification of SURF4 fluorescence intensity. n = 17 or 18 cells per condition across three independent batches.", fig.scap="SURF4 localisation in protrudin-expressing neurons"}

SURF4_plot
```

```{r SLC38A1-plot, echo = F, fig.width=16, fig.height=16, fig.path='../figures/full', fig.cap="A-F) Representative images of SLC38A1 in control, WT, and active protrudin-expressing cell bodies (A-C, respectively) and axon terminals (D-F).  All scale bars are 20 \\textmu{}m, and D-F are 20x20 \\textmu{}m.  G-I) Quantification of SLC38A1 fluorescence intensity. n = 15 to 18 cells per condition across three independent batches.", fig.scap="SLC38A1 localisation in protrudin-expressing neurons"}

SLC38A1_plot
```

```{r ERGIC53-plot, echo = F, fig.width=16, fig.height=16, fig.path='../figures/full', fig.cap="A-F) Representative images of ERGIC53 in control, WT, and active protrudin-expressing cell bodies (A-C, respectively) and axon terminals (D-F).  All scale bars are 20 \\textmu{}m, and D-F are 20x20 \\textmu{}m.  G-I) Quantification of ERGIC53 fluorescence intensity. n = 16 or 18 cells per condition across three independent batches.", fig.scap="ERGIC53 localisation in protrudin-expressing neurons"}

ERGIC53_plot
```

### Protrudin's effect on Golgi satellites

Given protrudin's effect on the ERGIC compartment, we were particularly interested in seeing if this also impacted axonal Golgi satellites.  As these are highly dynamic organelles, we decided to look at Golgi satellite transport in the presence of protrudin.  We co-expressed mCherry-tagged protrudin constructs with St3Gal5-EGFP, as a marker of Golgi satellites.  

```{r St3Gal-full-figure, echo = F, fig.width=16, fig.height=8, fig.path='../figures/full/', fig.cap="A) Representative axon image and kymographs showing St3Gal5 transport in the distal axon of cortical neurons expressing either mCherry (control) or mCherry-tagged protrudin constructs.  B) Total number of St3Gal5-labelled organelles in each cell imaged.  C) Average run length (per cell) of moving Golgi satellites  D) Average Golgi satellite velocity (per cell).  n = 22 or 24 cells per condition across three independent batches.", fig.scap="Protrudin effect on Golgi satellite (St3Gal5) transport in the distal axon"}

St3Gal5_kymo_plot
```

Following live imaging, these cells were fixed, and imaged at higher exposure on a confocal microscope to more clearly visualise Golgi satellites in the axon.  We measured both the number of puncta, as well as point intensity of each, as a proxy for size.  This confirmed that protrudin overexpression did not affect the overall number of Golgi satellites in the distal axon.

```{r GS-puncta-full-figure, echo = F, fig.width=16, fig.height=8, fig.path='../figures/full/', fig.cap="A) Representative axon image showing St3Gal5 puncta.  Scale bar is 20 \\textmu{}m.  B) Number of puncta per length of axon imaged.  C)Point fluorescence intensity of each St3Gal3 puncta counted.  n = 18 cells per condition across three independent batches.", fig.scap="Protrudin effect on Golgi satellite abundance and size in the distal axon"}

GS_puncta_full_figure
```

We then conducted a scratch assay to determine the localisation of Golgi satellites in the growth cones of mature neurons and examine whether protrudin might have an effect on axonal Golgi satellites in response to injury.  Cortical neurons cultured on glass coverslips were scratched with a sterile needle, then incubated for 18-20 hours before fixing.  Axons growing into the scratched region were imaged on a confocal microscope.

This revealed a strong concentration of St3Gal5 at the base of each growth cone, with slightly weaker signal extending further distally.  Where St3Gal5 was co-expressed with protrudin, it is apparent that the Golgi satellite signal reaches further into the growth cone than the ER.  This suggests that there are distinct ER and Golgi compartments in the growth cone, with secretory machinery oriented towards the direction of growth.  The images also suggest that there might be an accumulation of St3Gal5 immediately proximal to the growth cone in protrudin-expressing neurons, although further work is required to confirm and quantify this result.

```{r scratch-assay-full, echo = F, fig.width=16, fig.height=14, fig.path='../figures/full/', fig.cap = "Representative images of growth cones on DIV15 cortical neurons after scratch.  Neurons are co-expressing either mCherry (A) or mCherry-protrudin (B-C) with St3Gal5-EGFP.  Scale bars are 20 \\textmu{}m.  D-F) 20x20 \\textmu{}m images highlighting the growth cone details shown in A-C.  Arrows indicate large Golgi satellites, as well as the distal edge of protrudin localisation.  Note how the St3Gal5 signal extends beyond protrudin in these growth cones.", fig.scap="Scratch assay of neurons co-expressing protrudin and St3Gal5"}

scratch_fig
```

## Discussion

These results provide an insight into how protrudin might regulate local secretory pathways in the axon.  While protrudin does not appear to change the distribution of primarily ERES-associated proteins, such as BCAP31 and SURF4, active protrudin does increase the relative level of ERGIC53 in axon terminals.  As the ERGIC is a distinct compartment from the ER, this suggests a secretory role for protrudin beyond its effects on the axonal ER.  It is not yet clear if this represents an increased formation of ERGIC from ERES-derived vesicles in the axon, or increased transport of these structures from elsewhere in the cell.  Further work will use a GFP-tagged ERGIC53 construct to examine whether the ERGIC compartments are transported axonally, and whether this transport is affected by protrudin.

This is further supported by our finding that active protrudin affects Golgi satellite transport in the distal axon.  Unexpectedly, we found slower transport of Golgi satellites in the distal axon, particularly in the anterograde direction, despite unchanged run length or overall number of organelles.  Further experiments are required to identify a mechanism for this change, although it could represent a sustained interaction with another organelle during transport, potentially ERGIC or a related compartment.

These findings indicate that protrudin overexpression, and in particular active protrudin, can impact the local secretory machinery present in the axon, which may play a role in mediating protrudin's neuroprotective and regenerative effects.  Additionally, these results may account for the finding that active protrudin has a stronger regenerative effect than WT protrudin, but that the two constructs do not significantly differ in measures of Rab11 or integrin transport [@petrovaProtrudinFunctionsEndoplasmic2020].
