---
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
  word_document: default
bibliography: "`r here::here('bib/THESIS.bib')`"
always_allow_html: true
---

# Introduction {.unnumbered #INTRODUCTION}

```{=latex}
\markboth{INTRODUCTION}{}
```

```{r intro-setup, include = F}
knitr::opts_chunk$set(echo = FALSE)
source(here::here('style/thesis_setup.R'))

library(here)
library(readbitmap)

```

## The challenge of CNS regeneration

One of the greatest challenges of modern medicine is how to promote regeneration of the central nervous system (CNS).  Compared to most other tissues and organs in the human body, the CNS has exceptionally low regenerative ability, largely due to the neurons' inability to reliably regrow injured axons.  As a result, damage to the CNS — whether through trauma, infection, or disease — often results in lifelong disability.  In contrast, the peripheral nervous system (PNS) is capable of substantial regeneration after injury, with severed nerves able to regrow several centimetres, reach their targets, and form functional connections, suggesting that regeneration of the CNS might be medically possible, if we can identify the relevant factors at play.  While there has been active research in this area for decades (if not centuries), we are still a long way from what most people would consider 'successful' regeneration of damaged CNS tissues — the regrowth of damaged neuronal processes, appropriate remodelling of the extracellular environment, and restoration of functional synaptic connections.  In this section, I will summarise our current understanding of some of the factors limiting CNS regeneration, as well as the particular challenges I have decided to address with this project.

### Glaucoma

Glaucoma is the leading cause of irreversible sight loss worldwide.  It is estimated that over 3.5% of the world population aged 40-80 is affected by glaucoma, which has an enormous impact on quality of life for millions of people [@thamGlobalPrevalenceGlaucoma2014].  While there are treatments that can slow or even halt the progression of this disease, they cannot reverse damage that has already occurred, and the typical progression of glaucoma — with sight loss beginning at the periphery — means that it is often diagnosed at a late stage.  Furthermore, a significant percentage of patients do not respond to the first line treatment drugs, so there is a continued need for improved therapies [@rossettiEvaluationRateNonresponders2006; @ikedaLatanoprostNonrespondersOpenangle2006].

The deterioration of vision in glaucoma is associated with both the loss of axons in the optic nerve — which conveys visual information from the eye to the brain — and with the loss of retinal ganglion cell (RGC) bodies in the retina.  While many approaches so far have targeted the damaging increase in intraocular pressure typically considered a hallmark of glaucoma, there are cases where RGC degeneration occurs despite normal pressure in the eye (normal tension glaucoma), and it is not entirely clear what drives the disease progression in these cases [@killerNormalTensionGlaucoma2018].  Therefore, it is important to develop therapies that can target the neurodegeneration directly, in addition to those aimed at lowering intraocular pressure to prevent RGC damage.  

This combination of axon damage and neurodegeneration makes glaucoma an ideal target for CNS regeneration research, as any successful treatments have immediate medical relevance, as well as the potential to apply more broadly to a range of other CNS conditions.  To this end, we have been working to identify new gene therapy targets that can prevent the loss of RGCs and promote regeneration of RGC axons that have already been damaged.  By delving into the biology of why CNS axons fail to regenerate and how we can change this, we hope to introduce a new class of therapeutics that can be used to treat not only glaucoma, but also the numerous other CNS diseases and injuries that cause axonal damage or loss.

```{r glaucoma-schematic, echo = F, fig.width = 16, fig.height = 7, fig.cap = "Schematic showing normal human eye anatomy (left), as well as the changes caused by glaucoma (right).  In the normal eye, visual signals are transduced by the axons of retinal ganglion cells, which follow the arrows in the figure from the retina to the brain through the optic nerve.  In glaucoma, these signals are disrupted as axons in the optic nerve are damaged.", fig.scap="Glaucoma schematic"}
image_as_ggplot(read.bitmap(here('figures/external/glaucoma_horizontal.png')))
```


## Factors limiting CNS regeneration

The question of why CNS axons fail to regenerate after injury is a complex one, and our current understanding points to a wide range of both extrinsic and intrinsic factors, which interact to limit regrowth.^[Note that this regeneration is distinct from — but closely related to — *functional* recovery, which can result from reorganisation of surviving neuronal circuits after injury in the absence of any axonal regeneration.  The central nervous system is exceptionally flexible, and significant functional recovery can achieved after, for example, spinal cord injury, in the case of an incomplete injury and intensive rehabilitation.  However, this plasticity can be significantly enhanced by even limited axonal regeneration, if regrowing axons can bridge the site of an injury and create new connections between cells, even if the axon fails to regrow as far as its original targets.  In this thesis I will be focusing on survival and regrowth of damaged axons, as opposed to functional recovery after a CNS injury or disease.]

### Extrinsic factors

In 1980, Peter Richardson and colleagues at McGill showed that axons in a severed rat spinal cord could regenerate into an implanted sciatic nerve graft, thus demonstrating that CNS neurons were at least somewhat capable of regeneration, but were inhibited by their environment [@richardsonAxonsCNSNeurones1980].  Since then, researchers have identified a variety of environmental factors that limit CNS axon growth, ranging from differences in the cellular composition of PNS and CNS tissues, to changes in the extracellular environment and cell surface receptors that occur during development as the tissue matures.


#### Extraellular matrix

The extracellular environment of the mature central nervous system is unique from other tissues, and contains a number of chemicals that are known to be inhibitory to axonal growth, either mechanically or biochemically.  The composition of this environment changes during development, and becomes increasingly hostile to neurite growth as the CNS matures [@shewanExtensiveRegenerationVitro1995].  Many of these changes — in particular the secretion of perineuronal nets, matrix structures which surround certain types of neuron in the CNS — play an important role in stabilising and supporting synapses, and signal a change in CNS state from supporting growth to supporting electrical signalling [@kwokExtracellularMatrixPerineuronal2011; @fawcettRolesPerineuronalNets2019].  These developmental events also result in changes in the mechanical properties of CNS tissue, which can inhibit normal axon growth even in the absence of biochemical cues [@koserMechanosensingCriticalAxon2016a].

Given this substantial impact on neurite growth, numerous therapies tageting the extracellular matrix have been tested to promote regeneration after spinal cord and other CNS injuries, with mixed results.  It is now widely accepted that targeting inhibitory molecules in the extracellular matrix — either by emzymatic breakdown, as in the case of chondroitinase ABC, or by inhibiting synthesis, for example with 4-methylumbelliferone — can permit short-range regeneration and enhance neuronal plasticity [@bradburyChondroitinaseABCPromotes2002a; @@irvine4methylumbelliferoneEnhancesNeuroplasticity2023;].  These treatments have led to moderate restoration of function in some models of spinal cord injury, often through the reorganisation of local neuronal networks following an incomplete injury.  However, robust long-range axonal regeneration — past the site of injury and back to their original downstream targets — remains elusive [reviewed in @kwokExtracellularMatrixPerineuronal2011; @muirRecentAdvancesTherapeutic2019].

#### Cellular environment

Axonal regeneration, or its failure, is significantly influenced by surrounding non-neuronal cells.  In most cases, glia make up the vast majority of these cells, but in the case of traumatic injury to the CNS circulating immune cells can also play an important role.

##### Glia

One of the most obvious tissue-level differences between the CNS and the PNS is the type of glial cells present.  PNS axons are myelinated by Schwann cells, while in the CNS oligodendrocytes carry out the same role.  These cells have significantly different structures, and respond differently to injury.  Following an axonal injury, Schwann cells transform to exhibit a regeneration-promoting phenotype. In this state, they actively participate in clearing myelin debris, form a scaffold for axonal regrowth, and release exosomes containing growth-promoting factors [reviewed in @jessenRepairSchwannCell2016]. In contrast, oligodendroytes that survive an axonal injury in the CNS are largely quiescent, and do not appear to support axon repair in the same way [@ludwinOligodendrocyteSurvivalWallerian1990].  On the contrary, the myelin debris from damaged oligodendrocytes, both lipids and proteins, and in particular myelin-associated glycoprotein (MAG), are inhibitory to axon growth, and can contribute to inflammation following an injury to the CNS [@marMyelinLipidsInhibit2016; @mukhopadhyayNovelRoleMyelinassociated1994; @mckerracherIdentificationMyelinassociatedGlycoprotein1994].

##### Inflammation

In normal circumstances, microglia act as the primary immune cells of the central nervous system, while other immune cells are excluded by the blood-brain or blood-spinal cord barrier.  However, in the case of injury — or certain degenerative conditions, such as multiple sclerosis — circulating immune cells can infiltrate the CNS, causing significant inflammation [reviewed in @hellenbrandInflammationSpinalCord2021; @freyermuth-trujilloInflammationTargetTreatment2022; @grajchenPhysiologyFoamyPhagocytes2018].  While some transient inflammation appears to be important for clearing tissue debris and subsequent repair, this process commonly becomes dysregulated, and unresolved inflammation is a significant contributor to the secondary injury cascade seen after traumatic brain or spinal cord injury.  While the inflammatory changes that occur following CNS injury are highly complex and dynamic, this is an active area of research, and some studies have suggested that anti-inflammatory treatments could promote axon regeneration [@fuNonsteroidalAntiInflammatoryDrugs2007; @parkLocalImmunomodulationAntiinflammatory2018].

#### Synapses - a bridge between extrinsic and intrinsic factors

Recent work has increasingly pointed to the formation of synapses as a key point at which developing CNS neurons lose their intrinsic capacity to regenerate.   During this process, interactions with the external cellular environment trigger internal changes that encourage signalling function at the expense of inhibiting growth.  For example, Tedeschi et al. showed that a voltage-gated calcium channel subunit prevents axon growth, and that inhibiting the channel can promote regeneration *in vivo* [@tedeschiCalciumChannelSubunit2016].  Hilton et al. found a similar effect with presynaptic vesicle-tethering proteins, and also found that cells that were capable of regeneration down-regulated these proteins during axon growth [@hiltonActiveVesiclePriming2022].  (NOTE PNNs HERE?)  These results suggest that the cellular requirements to support synaptic signalling might be at odds with the mechanisms needed to support axonal growth, and that changing a neuron's interactions with external partners might be a necessary step in promoting regeneration through internal mechanisms.

### Intrinsic factors

In addition to extrinsic factors, developmental changes *within* neurons also present a barrier to CNS regeneration.  As the cells mature from a growth-permissive state into synaptically active neurons, they undergo structural and functional changes that correlate with the loss of regenerative ability [@kosekiSelectiveRab11Transport2017].  Additionally, many neurodegenerative diseases are associated with genetic or molecular dysfunction within neurons, as opposed to the extracellular environment, highlighting the importance of internal factors in maintaining healthy, functional axons.  Understanding and addressing these intrinsic sources of dysfunction and barriers to regeneration is essential to developing successful therapies to protect and repair CNS axons.

#### Cytoskeletal regulation of axon growth

Neurons undergo dramatic morphological changes during development, eventually forming extremely long, thin, and complex structures that, in many cases, must be carefully maintained for the entire lifetime of the organism.  To make this possible, mature neurons have a dynamic but highly regulated cytoskeleton, particularly in the axon.  However, the structural requirements of the mature, signal-carrying axon are very different from a growing axon, and in the case of axon injury this can prevent successful regeneration.   

A key feature of the mature neuronal cytoskeleton is the bundle of polarised microtubules that forms the backbone of the axon.  This structure is maintained by proteins such as EFA6, which truncates growing microtubules that leave the bundle to reach the axon periphery.  However, this can also inhibit axon outgrowth, which might limit regeneration [@quEfa6ProtectsAxons2019].

This microtubule structure, and by extension the structure of the axon, is also maintained by a repeated pattern of actin rings arranged at intervals around the circumference of the axon [@xuActinSpectrinAssociated2013; @quPeriodicActinStructures2017].  In addition to stabilising the microtubule bundle, recent evidence shows that these rings can expand and contract, playing a role in the regulation and speed of axonal transport [@wangRadialContractilityActomyosin2020].

At the axon tip, a delicate balance of microtubule polymerisation and stabilisation promotes forward extension of the growth cone and axon elongation during development.  This can be partially restored in mature axons with the addition of microtubule-stabilising drugs, but the effect is highly dose-dependent, with higher doses instead inhibiting axon growth (reviewed in [@blanquieCytoskeletonDynamicsAxon2018]).  There is also a complex interaction between these growing microtubules and the actin filaments that define the structure of the growth cone periphery.  Actin filaments in the growth cone are constantly polymerising, being pulled towards the center of the growth cone, and depolymerising, in a process that is dynamically regulated to limit growth speed and allow the growth cone to change direction in response to environmental cues (reviewed in [@leiteActinDynamicsGrowth2021]).  Inhibiting proteins involved in this retrograde flow of actin filaments has been shown to increase axon growth even in the presence of otherwise inhibitory cues [@hurEngineeringNeuronalGrowth2011; @yuMyosinIIActivity2012].

#### Growth-related signalling

The internal changes that occur during neuronal maturation are regulated by distinct signalling pathways.  Several of these have been associated with the loss of regenerative ability, and have been targeted to increase axon regeneration.

##### PTEN and mTOR

One of the most-studied signalling pathways in the context of CNS repair is the PI3K/AKT/mTOR pathway.  The levels of PIP~3~, an important signalling lipid, are developmentally regulated in neurons [@nieuwenhuisPI3kinaseDelta2020].  PIP~3~ is produced from PIP~2~ by PI3K, while the reverse reaction is catalysed by PTEN.  During growth, a balance of both enzymes maintains sufficient PIP~3~ levels to promote growth via AKT kinase and downstream targets such as mTOR.  As neurons mature, however, expression levels of PI3K decline.  The result is a decline in PIP~3~, and subsequent decrease in AKT signalling.  Restoring PIP~3~ levels in mature neurons, either through PTEN deletion or PI3K expression, can strongly promote axon regeneration [@parkPromotingAxonRegeneration2008; @nieuwenhuisPI3kinaseDelta2020;&nbsp;reviewed in @nieuwenhuisPromotingAxonRegeneration2022].

##### Other signalling pathways

Numerous other pathways have also been implicated in the failure of regeneration in the CNS.  Smith et al. found that SOCS3 — a negative regulator of cytokine signalling — inhibited axon regeneration in mouse RGCs by blocking responses to external growth factors, and that deleting SOCS3 promoted regeneration [@smithSOCS3DeletionPromotes2009].  Qiu et al. showed that DRG regeneration into the CNS following a conditioning lesion was dependent on cAMP and PKA to overcome inhibitory signals from the extracellular environment [@qiuSpinalAxonRegeneration2002].  These pathways all overlap or interact to some degree, but it is still unclear how they may be differentially regulated in the case of axonal injury, or if this varies across cell types.

#### Epigenetic and transcriptional regulation of regeneration

Dorsal root ganglia (DRG) neurons are unique in that they extend axons into both the peripheral and central nervous systems from their location just outside the spinal cord.  As peripheral neurons they have the intrinsic capacity for axonal regeneration, although this is usually only successful in the peripheral branch, due to the inhibitory extracellular environment of the spinal cord.  However, their regenerative capacity into the CNS is vastly enhanced by a prior injury to the peripheral axon, known as a conditioning lesion.  Changes in gene expression and regulation that occur in the weeks following a peripheral lesion can allow these cells to regenerate through the inhibitory environment of a spinal cord injury, raising the possibility that similar changes could convey similar capacity in neurons of the CNS [@neumannRegenerationDorsalColumn1999].  

This conditioning lesion phenomenon has provided an invaluable model to identifying epigenetic and transcriptomic changes that permit PNS regeneration but are normally lacking in the CNS.  For example, Puttagunta et al. found that histone modification was essential for the conditioning lesion effect, and activated signalling cascades to promote axonal growth [@puttaguntaPCAFdependentEpigeneticChanges2014].  Weng et al. found that DRG neurons upregulate TET3 after axon injury, which in turn demethylates DNA and allows the activation of regeneration-associated genes.  They also found that the regeneration induced by PTEN deletion in RGCs was dependent on TET1, a similar protein [@wengIntrinsicEpigeneticBarrier2017].  These studies highlight that chromatin accessibility and DNA modification are limiting factors associated with regenerative ability.  

Other work has identified transcription factors that may play a role in axon regeneration (reviewed in [@mooreMultipleTranscriptionFactor2011]).  Moore et al. suggest that different members of the KLF family of transcription factors differentially regulate the regenerative capacity of RGCs, while Ewan et al. pointed to ATF3 and c-Jun as key regulators [@mooreKLFFamilyMembers2009; @ewanAscendingDorsalColumn2021].  This is by no means an exhaustive list, and a significant amount of work is ongoing in this area to identify transcription factors that can be targeted to promote regeneration, particularly in combination with therapies that address chromatin accessibility (reviewed in [@wahaneEpigeneticRegulationAxon2019]).  

#### Changes in axonal transport with development

After synapse formation, axons undergo dramatic changes in transport, as growth machinery is excluded from the axon in favour of synapse components and synaptic vesicle precursors.  This results in decreased transport of some essential organelles, which can inhibit survival and regrowth in the case of injury.

##### Recycling endosomes

Among the organelles excluded from mature axons are recycling endosomes, which are marked with a small GTPase called Rab11.  Rab11 endosomes play a key role in recycling internalised receptors back to the plasma membrane, but they also play a role in transporting cargo into axons [@kosekiSelectiveRab11Transport2017].  One such cargo is integrins, which are essential for focal adhesion formation and involved with both developmental and regenerative axonal growth.  In DRG neurons in the PNS, integrin \textalpha{}9 can promote axon regeneration even into the inhibitory environment of the spinal cord [@cheahExpressionActivatedIntegrin2016].  However, integrins are excluded from mature CNS axons by another recycling endosome small GTPase called ARF6, in combination with its activator EFA6.  Inhibiting EFA6 allows integrins into the axon and promotes regeneration after axotomy *in vitro*, but it remains to be seen whether this approach is effective *in vivo* [@evaEFA6RegulatesSelective2017].

##### Mitochondria

Similar changes in mitochondrial motility have been identified.  Lewis et al. found that mitochondrial motility decreased with age in cortical neurons, and mature mitochondria were often stationary near presynapses  [@lewisProgressiveDecreaseMitochondrial2016].   Sainath et al. found that chondroitin sulphate proteoglycans — potent inhibitors of axon regeneration — prevented mitochondria from reaching the growth cones of regenerating DRG neurons [@sainathChondroitinSulfateProteoglycans2017].  Other studies linked this change even more directly to axon regeneration, showing that artificially enhancing mitochondrial motility by either reducing the anchor protein syntaphilin or overexpressing transport-regulating ARMCX1 could boost regeneration after axon injury [@zhouFacilitationAxonRegeneration2016; @cartoniMammalianSpecificProteinArmcx12016].  

#### The virtuous cycle of axon growth

Many of these intrinsic factors interact in what has been described as the 'virtuous cycle of axon growth' [@petrovaVirtuousCycleAxon2018].  Axonal growth is sustained by a cycle in which activated axonal growth factor receptors trigger transcriptional and signalling cascades through retrograde transport to the cell body.  These in turn can result in the translation and subsequent delivery of more receptors and growth machinery, which allow the cell to respond more effectively to growth cues in the extracellular environment.  As the cell matures, changes in axon transport lead to lower levels of growth factor receptors in the distal axon, and the cycle stops.  By targeting any stage of this cycle — axon transport, signalling, transcription, or translation — it should be possible to effectively restart the cycle and promote regeneration.

In summary, cytoskeletal organisation, signalling pathways, epigenetic and transcriptional regulation, and organelle transport have all been identified as key intrinsic barriers to axonal regeneration.  The virtuous cycle theory highlights links between these systems, and suggests that growth-promoting interventions have the potential to feed forward to dramatically enhance axonal regeneration.

## Protrudin {#intro-ptdn}

Protrudin is an interesting protein in the context of axon regeneration because it sits at the nexus of several of these intrinsic systems that can prevent or promote regeneration.  As a result, targeting protrudin has the potential to substantially influence a neuron's regenerative capacity by synergistically acting on multiple different pathways.

### Structure and localisation

```{r ptdn-schematic, echo = F, fig.height = 4, fig.cap = "Schematic of protrudin embedded in the ER membrane, showing RBD (Rab binding domain), transmembrane domains, FFAT motif (two phenylalanines in an acidic tract), CC motif (coiled-coil), and FYVE domain (Fab1 YOTB Vac1 EEA1 zinc finger domain).  The ER lumen is to the bottom of the image.", fig.scap="Protrudin schematic"}
image_as_ggplot(read.bitmap(here('figures/external/protrudin_schematic.png')))
```

Protrudin is a transmembrane ER protein.  It contains an intramembrane hairpin loop, which restricts its localisation to tubular domains of the ER, where it contributes to shaping and maintaining membrane curvature [@hashimotoProtrudinRegulatesEndoplasmic2014].  This loop is also required for protrudin oligomerisation - functional protrudin forms homodimers or tetramers, although how this contributes to its function is not well understood [@pantakaniOligomerizationZFYVE27Protrudin2011].  Protrudin also contains an array of binding domains exposed to the cytoplasm, which include a Rab-binding domain (RBD), an FFAT domain, which interacts with VAPA/B at ER contact sites, a coiled-coil (CC) domain, which binds kinesin, and a FYVE domain, which interacts with phospholipids.  There is also a predicted low-complexity region between the hairpin and CC domains, which has been shown to mediate interaction with Rab7 [@raiborgRepeatedEREndosome2015].  

### Cellular outgrowth, migration, and invasion

Protrudin, otherwise known as ZFYVE27, is so named because its overexpression induces the growth of long cellular protrusions in HeLa cells, which are non-neuronal cells derived from an epithelial tumour [@shiraneProtrudinInducesNeurite2006].  Subsequent work has shown similar effects on a number of cell types [@zhangRoleSpastinProtrudin2012], and indicated a role for protrudin in outgrowth-dependent processes, such as migration and cancer cell invasion [@aroraProtrudinRegulatesFAK2022; @pedersenProtrudinmediatedEREndosome2020].  However, the underlying mechanisms of these effects are not fully understood, and there are a number of potential pathways that could contribute.

### Transport{#intro-ptdn-transport}

Protrudin contains a kinesin binding site, indicating a role in mediating motor protein-mediated transport.  In particular, this would promote plus-end directed transport, which is essential for delivering cargo into the long axons of CNS neurons.  Matsuzaki et al. studied this in some depth, and found that protrudin's outgrowth-promoting effect in HeLa cells is attenuated if KIF5 is inhibited, and increased if KIF5 is also overexpressed [@matsuzakiProtrudinServesAdaptor2011].  They found evidence for direct protrudin-KIF5 interaction in mouse brains, and identified protrudin as the KIF5 adaptor for SURF4, Rab11, VAPA/B, and RTN3.  As these proteins are essential for ER function, contact site formation, protein transport, and a number of other processes, this suggests that, even at the low levels found endogenously in CNS neurons [@kosekiSelectiveRab11Transport2017; @petrovaProtrudinFunctionsEndoplasmic2020], protrudin might play an important role in normal axonal function.

The study by Petrova et al. [@petrovaProtrudinFunctionsEndoplasmic2020], which I contributed to during the first year of my PhD, looked at protrudin specifically in the context of integrin transport.  In particular, integrins \textalpha{}9 and \textbeta{}1, which are capable of promoting axon growth in the mature spinal cord [@cheahExpressionActivatedIntegrin2016], are cargo of Rab11 endosomes [@evaRab11ItsEffector2010].  We found that protrudin overexpression had a small effect on developmental neurite outgrowth, but strongly promoted axon regrowth after axotomy *in vitro*.  In addition, protrudin prevented retinal ganglion cell death after an axon crush injury *in vivo*, and allowed axons to regrow past the injury site.  I discuss this result further in Chapter 1, section \@ref(protrudin-ONC).

In contrast to our work on Rab11 and recycling endosomes, Raiborg et al. looked at protrudin's effect on the transport of late endosomes, marked by the small GTPase Rab7.  Using PC12 cells, which are derived from a neuroendocrine tumour,  they found that protrudin *indirectly* promoted late endosome transport, forming transient ER-late endosome contacts at which a separate adaptor, FYCO1 (ZFYVE7), was able to bind Rab7 and KIF5 to move late endosomes down the axon [@raiborgRepeatedEREndosome2015].

### ER contact sites

Other studies have indicated a direct role for protrudin at ER-organelle membrane contact sites (MCS).  Protrudin interacts with VAPA/B, key tethers at many MCS [@saitaPromotionNeuriteExtension2009; @matsuzakiProtrudinServesAdaptor2011; @hashimotoProtrudinRegulatesEndoplasmic2014; @elbaz-alonPDZD8InteractsProtrudin2020].  It may also act at ER-mitochondrial contact sites, which are important for axonal function and growth [@elbaz-alonPDZD8InteractsProtrudin2020; @leeIncreasedERMitochondria2019], and may be disrupted in neurodegenerative disease (reviewed in [@bernard-marissalEndoplasmicReticulumMitochondria2018]).  In addition, protrudin itself has been shown to act as a tether at ER-endosome contact sites, as it contains a phosphoinositide-binding FYVE domain, as well as several motifs that can bind Rab proteins [@raiborgRepeatedEREndosome2015].  It may also play a role in ER-plasma membrane contacts, as protrudin contains a non-canonical FYVE domain that interacts more strongly with PM lipids than those more abundant on endosomes [@gilPhosphoinositidesDifferentiallyRegulate2012].  

### Axonal ER{#intro-ptdn-axonalER}

The axonal ER is typically comprised of only a few very narrow tubules, but is essential for normal axonal function [@terasakiAxonalEndoplasmicReticulum2018].  As protrudin localises to tubular ER, it can be found in the axon, although in mature neurons it is more abundant in the cell body and dendrites.  That being said, it does not appear to be actively excluded from the axon, and it readily moves into the axon when overexpressed [@petrovaProtrudinFunctionsEndoplasmic2020].  Protrudin is known to interact with a number of important axonal ER components, although it remains unclear what endogenous function protrudin has in this compartment.  

Protrudin contains a membrane hairpin domain, which is shared with other proteins that play a role in ER membrane shaping [@changProtrudinBindsAtlastins2013].  Several of these are implicated in hereditary spastic paraplegias (HSP), a group of axon degenerative diseases, which may stem from fragmentation of the axonal ER [@yalcinModelingAxonalEndoplasmic2017].  Protrudin itself has been considered an HSP gene [@mannanZFYVE27SPG33Novel2006], although there is some controversy in this area, and the 'mutation' involved may actually be a benign polymorphism [@martignoniRoleZFYVE27Protrudin2008].  However, protrudin does interact with spastin, atlastin, and other HSP proteins, so it may still play a role in these conditions [@changProtrudinBindsAtlastins2013].  

While dysfunction of these ER-shaping proteins can lead to axonal degeneration, their function is also required for axon regeneration, where they appear to concentrate ER tubules near the growing tip [@raoSpastinAtlastinER2016].  Protrudin may act in the same way, as Petrova et al. found that protrudin increases the amount of ER in growth cones [@petrovaProtrudinFunctionsEndoplasmic2020].  However, while these studies demonstrate that the axonal ER is involved in axon regeneration, it is not currently clear what functions it provides in this context.

The axonal ER also acts as a local calcium store, and its capacity is regulated by membrane shaping proteins via tubule diameter.  Protrudin may act in the same way, although it is unclear how this may or may not contribute to protrudin's regenerative function, as different ER-shaping proteins have been shown to have opposing effects on axon regeneration [@konnoEndoplasmicReticulumMorphological2021; @alhajlahOverexpressionReticulonEnhances2021].

### Previous work on protrudin

With these diverse roles across numerous cellular functions, protrudin could be influencing neuronal survival and regeneration through several different mechanisms.  Our lab has been working with protrudin for some time now, having initially become interested in the protein in the context of integrin transport.  This led to the publication of a paper which I contributed to [@petrovaProtrudinFunctionsEndoplasmic2020], some of which is presented in section \@ref(protrudin-ONC).

That paper presented a number of key findings, which I have built upon over the course of this project.  The foremost of these is that protrudin overexpression has a striking effect on neuronal survival and regeneration after axon injury.  

## Aims of this project{#intro-aims}

My initial aim in starting this project was to further examine the effect of protrudin on *in vivo* CNS injury, with the intent of developing it towards a gene therapy.  Unfortunately, the 2020 coronavirus pandemic and associated disruption necessitated a dramatic change in direction, and I have subsequently focused on understanding which cellular pathways contribute to protrudin's beneficial effects.  This is primarily intended to inform the translation of protrudin as a gene therapy for neurodegenerative conditions and CNS injury, although it also lays the groundwork for identifying new therapeutic targets in the same pathways.
